### 2017-09-02 paper/main.tex
add latex for paper

\documentclass{acm_proc_article-sp}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{tabularx}

% solution for getting around this issue to use the titlesec package:
% https://tex.stackexchange.com/questions/110546/spacing-below-section-title
\makeatletter
\newcommand\subparagraph{%
  \@startsection{subparagraph}{5}
  {\parindent}
  {3.25ex \@plus 1ex \@minus .2ex}
  {-1em}
  {\normalfont\normalsize\bfseries}}
\makeatother
\usepackage{titlesec}
\let\subparagraph\relax % You don't want to use \subparagraph

\begin{document}

\titlespacing\section{0pt}{8pt plus 4pt minus 2pt}{5pt plus 1pt minus 1pt}
\titlespacing\subsection{0pt}{5pt plus 4pt minus 2pt}{5pt plus 1pt minus 1pt}

% define code block framing style
\BeforeBeginEnvironment{minted}{
  \begin{mdframed}[skipabove=15pt,
                   skipbelow=10pt,
                   linewidth=0pt,
                   innerleftmargin=5pt,
                   innertopmargin=0pt,
                   innerbottommargin=0pt]}
\AfterEndEnvironment{minted}{\end{mdframed}}

\conferenceinfo{}{Bloomberg Data for Good Exchange 2016, NY, USA}

\title{
  Themis-ml:
}
\subtitle{
    A Fairness-aware Machine Learning Interface for \\
    End-to-end Discrimination Discovery and Mitigation
}

\numberofauthors{1}
\author{
\alignauthor
Niels Bantilan\\
       \affaddr{Arena.io}\\
       \affaddr{New York, NY}\\
       \email{niels.bantilan@gmail.com}
}

\maketitle

\begin{abstract}
As more industries integrate machine learning into socially sensitive decisions
processes like hiring, loan-approval, and parole-granting, we are at risk of
perpetuating historical and contemporary socioeconomic disparities. This is a
critical problem because on the one hand, organizations who use but do not
understand the discriminatory potential of such systems will facilitate the
widening of social disparities under the assumption that algorithms are
categorically objective. On the other hand, the responsible use of machine
learning can help us measure, understand, and mitigate the implicit historical
biases in socially sensitive data by expressing implicit decision-making mental
models in terms of explicit statistical models. In this paper we specify,
implement, and evaluate a ``fairness-aware'' machine learning interface called
themis-ml, which is intended for use by individual data scientists and
engineers, academic research teams, or larger product teams that use machine
learning in production systems.
\end{abstract}

% A category with the (minimum) three required fields
% \category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

%\category{H.4}{Information Systems Applications}{Miscellaneous TODO}

%\terms{TODO}

%\keywords{TODO}

\section{Introduction} In recent years, the transformative potential of machine
learning (ML) in many industries has propelled ML into the forefront of
mainstream media. From improving products and services to optimizing logistics
and operations, ML and artificial intelligence more broadly offer a wide range
of tools for organizations to enhance their internal and external capabilities.

As with any tool, we can use ML to engender great social benefit, but as
\cite{o2017weapons} emphasizes, we can also misuse it to bring about devastating
harm. In this paper, we focus on ML systems in the context of Decision Support
Systems (DSS), which are software systems that are intended to assist humans in
various decision-making contexts \cite{yoshimura2006decision,
montgomery2000evaluation, barnett1987dxplain, mysiak2005towards}. The misuse of
ML in these types of systems could potentially precipitate a widespread adverse
impact on society by introducing insidious feedback loops between biased
historical data and current decision- making \cite{o2017weapons}.

Researchers have developed many discrimination discovery and fairness-aware ML
methods \cite{kusner2017counterfactual, kamiran2012data, kamishima2012fairness,
kamiran2012decision, zemel2013learning, zafar2017fairness, dwork2012fairness,
zliobaite2015survey}, so we build on work done by others and seek to leverage
these techniques in the context of research- and product-based machine learning
applications. Our contributions in this paper are two-fold. First, we propose an
application programming interface (API) for ``Fairness-aware Machine Learning
Interfaces'' (FMLI) in the context of a simple binary classifier. Second, we
introduce themis-ml, an FMLI-compliant library, and apply it to a hypothetical
loan-granting DSS using the German Credit Dataset \cite{bache2013uci}. Our hope
is that themis-ml serves as a reference implementation that others might use and
extend for their own purposes.

%\newpage
\section{Bias and Discrimination}
Colloquially, bias is simply a preference for or against something, e.g.,
preferring for vanilla over chocolate ice cream. While this definition is
intuitive, there are two types of bias that are less obvious and important to
highlight in this paper: algorithmic and machine learning bias.

Algorithmic bias is when a set of mathematical rules favors one set of
attributes over others in relation to some target variable, like ``approve'' or
``deny'' a loan. Machine learning bias is when a trained model systematically
generates predictions that favor one group over another in relation to some
set of attributes, e.g., education, and some target variable, e.g., ``default
on credit''. While these definitions of bias are amoral, discrimination is in
essence moral, occurring when an action is based on biases resulting in the
unfair treatment of people. We define fairness as the inverse of
discrimination, meaning that a ``fairness-aware'' model is one that produces
non-discriminatory predictions.

Bias can lead to either direct (intended/explicit) or indirect
(unintended/implicit) discrimination, and the predominant legal concepts used to
determine these two types are known as disparate treatment and disparate impact
respectively \cite{barocas2016big}. As \cite{kusner2017counterfactual,
kamiran2012data} suggest, we can address disparate treatment in ML models by
simply removing all variables that are highly correlated to the protected class
of interest, in addition to the protected class itself, from the training data.
However, as \cite{kusner2017counterfactual} points out, doing so does not
necessarily mitigate discriminatory predictions and may actually introduce
unfairness into an otherwise fair system. In contrast, addressing disparate
impact is more complex because it depends on historical processes that generated
the training data, non-linear relationships between the features and protected
class, and whether we are interested in measuring individual- or group-level
discrimination \cite{dwork2012fairness}.

\section{A Fairness-aware Machine Learning Interface} So how does one measure
disparate impact and individual-/group-level discrimination in an ML-driven
product? In this section, we describe the main components of a simple
classification system, enumerate a few of the use cases that a research or
product team might have for using an FMLI, and propose an API that fulfills
these use cases.

A simple classification ML pipeline consists of five steps: data ingestion, data
preprocessing, model training, model evaluation, and prediction generation on
new e\(X_a\)mples. Data ingestion is outside the scope of this paper because it is a
highly variable process that depends on the application, often involves
considerable engineering effort, and potentially requires external stakeholder
bu\(y^{-})\)in.

Table 1 below outlines a simple classification system in terms of the core
interfaces in scikit-learn (sklearn), which is a machine learning library in the
Python programming language \cite{buitinck2013api}, and Table 2 delineates some
of the use cases that research or product teams might have to justify the use of
an FMLI.

\begin{table}
  \caption{A Simple Classification Pipeline}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l X X}
    \textbf{API Interface} & \textbf{Function} & \textbf{E\(X_a\)mples} \\
    \hline
    \textbf{Transformer} &
      \emph{Preprocess} raw data for model training. &
      mean-unit variance scaling, min-max scaling \\
    \textbf{Estimator} &
      \emph{Train} models to perform a classification task. &
      logistic regression, random forest \\
    \textbf{Scorer} &
      \emph{Evaluate} performance of different models. &
      accuracy, f1-score, area under the curve \\
    \textbf{Predictor} &
      \emph{Predict} outcomes for new data. &
      single-classifier prediction, ensemble prediction
  \end{tabularx}
\end{table}

\section{FMLI Specification} Here we propose a high-level specification of
themis-ml, an FMLI named after the ancient Greek titaness of justice. We adopt
sklearn's principles of consistency, inspection, non-proliferation of classes,
composition, and sensible defaults \cite{buitinck2013api}, and extend them with
the following FMLI-specific principles:

\begin{hangparas}{.15in}{1}
  \textbf{Model flexibility.} Focus on fairness-aware methods that are applicable
    to a variety of model types because users might have no control or full
    control over the specific model training implementation.
\end{hangparas}
\begin{hangparas}{.15in}{1}
  \textbf{Fairness as performance.} Provide estimators and scoring metrics that
    explicitly encode a notion of both model accuracy and fairness so that models
    can optimize for both.
\end{hangparas}
\begin{hangparas}{.15in}{1}
\textbf{Transparency of fairness-utility tradeoff.} Fair models often make less
  accurate predictions [8, 13], which is an important factor when assessing
  their business impact.
\end{hangparas}

\begin{table}
  \caption{FMLI Use Cases}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{X X}
    \textbf{Use Case} & \textbf{Rationale} \\
    \hline
    Detect and reduce discrimination in a production machine learning pipeline.
      & Fairness-aware modeling aligns with team/company values,
        provides protection from legal liability. \\
    Measure individual-/group-level discrimination in data with respect to a
    protected class and outcome of interest.
      & Need to assess the potential bias resulting from training models on
        data. \\
    Preprocess raw data or post-process model predictions in a way that reduces
    discriminatory predictions generated by models.
      & Unable to change the underlying implementation of the model training
        process. \\
    Explicitly learn model parameters that produce fair predictions for a variety
    of model types.
      & Need for flexibility when experimenting with or deploying different model
        types. \\
    Evaluate the degree to which fairness-aware methods reduce discrimination
    and assess the fairness-utility tradeoff.
      & Need for assessing the business consequences or other implications of
        deploying a fairness-aware model.
  \end{tabularx}
\end{table}

\subsection{Preliminaries} In the following subsections we describe specific
methods from the ML fairness literature that map onto each of the sklearn
interfaces. Note that we only provide a high level summary of each method,
citing the original sources for more implementation details. The following
descriptions make two assumptions: (i) the positive target label \(y^{+}\)
refers to a desirable outcome, e.g. ``approve loan'', and vice versa for the
negative target label \(y^{-}\), and (ii) the protected class is a binary
variable defined as \(s \in {d, a}\), where \(X_d\)  are members of the
disadvantaged group and \(X_a\) are members of the advantaged group.

Following these conventions, we define \(X_d\), \(y^{+}\) and \(X_d\), \(y^{-}\)
as the set of observations of the disadvantaged group that are positively
labelled and negatively labelled respectively. Similarly, \(X_a\), \(y^{+}\) and
\(X_a\), \(y^{-}\)  are observations of the advantaged group that are positively
and negatively labelled respectively.

\subsection{Transformer} The main idea behind fairness-aware preprocessing is to
take a dataset D consisting of a feature set \(X_train\), target labels
\(y_train\), and protected class strain to output a modified dataset.

Massaging modifies ytrain by relabelling the target variables in such a way that
``promotes'' members of the disadvantaged protected class (e.g. ``immigrant'')
and ``demotes'' members of the advantaged class (e.g. ``citizen'')
\cite{kamiran2012data}. A ranker model R (e.g. logistic regression) is trained
on D and ranks are generated for all observations. Some of the top-ranked
observations \(X_d\), \(y^{-}\)  are ``promoted'' to \(X_d\), \(y^{+}\) and some
of the bottom-ranked observations \(X_a\), \(y^{+}\)  are ``demoted'' to
\(X_a\), \(y^{-}\) such that the proportion of \(y^{+}\) are equal in both
\(X_d\) and \(X_a\). One caveat is that this method is intrusive because it
directly manipulates \(y\) such that it no longer reflects the original value
found in the raw data.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import Massage
  from sklearn.linear_model import LogisticRegression

  # use logistic regression as the ranking algorithm
  massager = Massage(ranker=LogisticRegression)

  # obtain a new set of labels
  y = massager.fit_transform(X, y, s)
\end{minted}

Reweighting takes a dataset D and assigns a weight to each observation using
conditional probabilities based on target labels and protected class membership
\cite{kamiran2012data}. In brief, large weights are assigned to \(X_d\),
\(y^{+}\) and \(X_a\), \(y^{-}\) , while small weights are assigned to \(X_d\),
\(y^{-}\) and \(X_a\), \(y^{+}\). The weights are then used as input to model
types that support weighted observations, which points to the main limitation of
this method, since not all classifiers can incorporate observation weights
during the learning process.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import Reweight

  reweigher = Reweight()

  # obtain fairness-aware weights for each observation
  reweigher.fit(y, s)
  weights = reweigher.transform(y, s)
\end{minted}

Sampling is composed of two methods: the first involves uniformly sampling n
observations from each group, where n is the expected size of that group
assuming a uniform distribution. The second is to preferentially sample
observations using a ranker R, similar to the massaging method. The procedure is
to duplicate the top-ranked \(X_d\), \(y^{+}\) and \(X_a\), \(y^{-}\) while
removing top-ranked \(X_d\), \(y^{-}\) and \(X_a\), \(y^{+}\)
\cite{kamiran2012data}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import (
      UniformSample, PreferentialSample)
  from sklearn.linear_model import LogisticRegression

  # use logistic regression as the ranking algorithm
  uniform_sampler = UniformSample()
  preferential_sampler = PreferentialSample(
      ranker=LogisticRegression)

  # obtain a new dataset with uniform sampling
  uniform_sampler.fit(y_train, s_train)
  X, y, s = uniform_sampler.transform(X, y, s)

  # obtain a new dataset with preferential sampling
  preferential_sampler.fit(y_train, s_train)
  X, y, s = preferential_sampler.transform(X, y, s)
\end{minted}

\subsection{Estimator}

Themis-ml implements two methods for training fairness-aware models: the
prejudice remover regularizer (PRR), and the additive counterfactually fair
(ACF) model.

\cite{kamishima2012fairness} proposes PRR as an optimization technique that
extends the standard L1/L2-norm regularization method \cite{ng2004feature,
ribeiro2016should} by adding a prejudice index term to the objective function.
This term is equivalent to normalized mutual information, which measures the
degree to which predictions \(y\) and s are dependent on each other. With values
ranging from 0 to 1, 0 means that \(y\) and s are independent and a value of 1
means that they are dependent. The goal of the objective function is to find
model parameters that minimize the difference between \(y\) and \(y\) in
addition to the degree to which \(y\) depends on s.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.linear_model import LogisticRegressionPRR

  # use L2-norm regularization and prejudice index as
  # the discrimination penalizer
  lr_prr = LogisticRegressionPRR(
      penalty="L2", discrimination_penalty="PI")

  # fit the models
  lr_prr.fit(X, y, s)
\end{minted}

ACF is a method described by \cite{kusner2017counterfactual} within the
framework of counterfactual fairness. The main idea is to train linear models to
predict each feature using the protected class attribute(s) as input. We can
then compute the residuals \(\epsilon_{ij}\) between the predicted feature
values and true feature values for each observation i for each feature j. The
final model is then trained on \(\epsilon_{ij}\) as features to predict y.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.linear_model import LinearACFClassifier

  # by default, LinearACFClassifier uses linear
  # regression as the continuous feature estimator
  # and logistic regression as the binary feature
  # estimator and target variable classifier
  linear_acf = LinearACFClassifier()

  # fit the models
  linear_acf.fit(X_train, y_train, s_train)
\end{minted}

\subsection{Predictor}

Themis-ml draws on two methods to make model type-agnostic predictions: Reject
Option Classification (ROC) and Discrimination Aware Ensemble Classification
(DAEC) \cite{kamiran2012decision}. Unlike the Transformer and Estimator methods
outlined above, ROC and DAEC do not modify the training data or the training
process. Rather, they postprocess predictions in a way that reduces
discriminatory predictions.

\cite{kamiran2012decision} describes two ways of implementing ROC, starting with
ROC in a single classifier setting. ROC works by training an initial classifier
on D, generating predicted probabilities on the test set, and then computing the
proximity of each prediction to the decision boundary learned by the classifier.
Within the critical region threshold \(\theta\) around the decision boundary,
where 0.5 < \(\theta\) < 1, \(X_d\) are assigned as \(y^{+}\) and \(X_a\) are
assigned as \(y^{-}\). ROC in the multiple classifier setting is similar to the
single classifier setting, except that predicted probabilities are defined as
the weighted average of probabilities generated by each classifier.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.postprocessing import (
      SingleROClassifier, MultiROClassifier)
  from sklearn.linear_model import LogisticRegression
  from sklearn.tree import DecisionTreeClassifier

  # use logistic regression for single classifier setting
  single_roc = SingleROClassifier(
      estimator=LogisticRegression())

  # use logistic regression and decision trees for
  # multiple classifier setting
  multi_roc = MultiROClassifier(
      estimators=[LogisticRegression(),
                  DecisionTreeClassifier()])

  # fit the models and generate predictions
  single_roc.fit(X, y, s)
  multi_roc.fit(X, y, s)
  single_roc.predict(X, s)
  multi_roc.predict(X, s)
\end{minted}

The main limitation of ROC is that model types must be able to produce predicted
probabilities. DAEC gets around this problem by training an ensemble of
classifiers and, through a similar relabelling rule as ROC, re-assigns any
prediction where classifiers disagree on the predicted label. As
\cite{kamiran2012decision} notes, in general, the larger the disagreement
between classifiers, the larger the reduction in discrimination.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.postprocessing import DAEnsembleClassifier
  from sklearn.linear_model import LogisticRegression
  from sklearn.tree import DecisionTreeClassifier

  # use logistic regression and decision trees
  dae_clf = DAEnsembleClassifier(
      estimators=[LogisticRegression(),
                  DecisionTreeClassifier()])

  # fit the models and generate predictions
  dae_clf.fit(X, y, s)
  dae_clf.predict(X, s)
\end{minted}

\subsection{Scorer}

The Scorer interface is concerned with measuring the degree to which data or
predictions are discriminatory. Themis-ml implements two methods for measuring
group-level discrimination and two methods for measuring individual-level
discrimination.

In the context of measuring group-level discrimination,
\cite{zliobaite2015survey} describes mean difference and normalized mean
difference. Mean difference measures the difference between \(p(a \cup y^{+})\)
and \(p(d \cup y^{+})\). Values range from -1 to 1, where -1 is the reverse-
discrimination case (all \(X_a\) have \(y^{-}\) labels and all \(X_d\) members
have \(y^{+}\) labels) and 1 is the fully discriminatory case (all \(X_a\) have
\(y^{+}\) labels and all \(X_d\) have \(y^{-}\) labels). Normalized mean
difference scales these values to 0 and 1 based on the maximum possible
discrimination in a dataset given the rate of positive labels
\cite{zliobaite2015survey}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.metrics import (
      mean_difference, normalized_mean_difference)

  # compare group-level discrimination in true
  # labels and predicted labels
  md_y_true = mean_difference(y, s)
  md_y_pred = mean_difference(pred, s)
  md_y_pred - md_y_true

  norm_md_y_true = norm_mean_difference(y, s)
  norm_md_y_pred = norm_mean_difference(pred, s)
  norm_md_y_pred - norm_md_y_true
\end{minted}

\cite{zliobaite2015survey} also describes consistency and situation test score
as individual-level discrimination measures. Consistency measures the difference
between the target label of a particular observation and target labels of its
neighbors. K-nearest neighbors (knn) measures the pairwise distance between
observations X. Then, for each observation \(x_i\) and each neighbor \(x_j \in
knn(x_i)\), we compute the differences between \(y_i\) and target labels of
neighbor \(y_i\). A consistency score of 0 indicates that there is no
individual-level discrimination and a score of 1 indicates that there is maximum
discrimination in the dataset.

The situation test score metric is similar to consistency, except we consider
only \(x_i \in X_d\). This method uses mean difference to compute a
discrimination score among neighbors \(x_j \in knn(x_i)\), producing a score
between 0 and 1, where 0 indicates no discrimination and 1 indicates maximum
discrimination \cite{zliobaite2015survey}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.metrics import (
      consistency, situation_test_score)

  # compare individual-level discrimination
  # in true labels and predicted labels
  c_true = consistency(y, s)
  c_pred = consistency(y, s)
  c_pred - c_true

  sts_true = situation_test_score(y, s)
  sts_pred = situation_test_score(y, s)
  sts_pred - sts_true
\end{minted}

\section{Evaluating Themis-ml}

\emph{\textbf{Work in progress}. Using the German Credit dataset, we evaluate
the efficacy of themis-ml in reducing discriminatory predictions. We do this by
comparing the performance of various fairness-aware techniques with a standard
``fairness- unaware'' classification system, as measured by the four
discrimination measures described above.}

\section{Discussion}

In this paper, we describe FMLIs in the classification context where we consider
only a single binary protected class variable. More work needs to be done to
generalize FMLIs to the multi-classification, regression, and multiple protected
classes contexts. Additionally, little is understood about the question of the
composability of fairness-aware methods. In other words, when different
techniques are used together in sequence, are the resulting discrimination
reductions additive or otherwise?

Future technical work might also extend the FMLI specification to include
techniques like Locally Interpretable Model-Agnostic Explanations
\cite{ribeiro2016should} and develop legal frameworks for thinking about how
different stakeholders would interact with FMLIs. Continued conversations among
players in civil society, industry, academia, and government is needed to
clearly articulate which parts of an FMLI should or should not be exposed to
particular stakeholders and why. For example, the model-training components
of the FMLI should not be accessible to external auditors for intellectual
property reasons, but perhaps they might have limited access to the predictions
generated by the models.

Finally, our ability to measure and mitigate discrimination is limited by our
common social, legal, and political understanding of discrimination itself. This
common understanding is often lacking because marginalized social groups
typically do not have a voice at the table when defining what counts and does
not count as discrimination. Therefore, we see the need for continued discussion
around the meaning of discrimination in different contexts. Since FMLIs are
simply a tool to measure and mitigate formalized definitions of discrimination,
it is important for all stakeholders to engage in an inclusive forum where
everyone, especially disadvantaged social groups, can contribute.

% \begin{figure}[h]
% \centering
% \includegraphics[width=4cm]{IMG/d4gx-2015-logo.png}
% \caption{Conference logo for 2015 (PNG)}
% \label{d4gx-logo-2015}
% \end{figure}

% \begin{figure*}
%   \includegraphics[width=\textwidth]{IMG/d4gx-2015-logo.pdf}
%   \caption{Same logo, full width.}
% \end{figure*}

% \begin{figure}[h]
% \centering
% \includegraphics[width=7cm]{IMG/d4gx-2015-logo.pdf}
% \caption{Conference logo for 2015 (PDF)}
% \label{d4gx-logo-2015}
% \end{figure}

\nocite{*}
\bibliographystyle{ieeetr}
\bibliography{references}

\end{document}

### 2017-09-02 paper/build.sh

@@ -1,0 +1,2 @@
+pdflatex -shell-escape main
+bibtex main

### 2017-09-06 paper/main.tex
complete draft of final paper submission

@@
 \usepackage{minted}
+\usepackage{subcaption}
 \usepackage{tabularx}
+\usepackage[labelfont=bf]{caption}
+\captionsetup{labelfont = bf}
 
@@
 \let\subparagraph\relax % You don't want to use \subparagraph
+
+\setlength{\parskip}{1em}
 
@@
 preferring for vanilla over chocolate ice cream. While this definition is
-intuitive, there are two types of bias that are less obvious and important to
-highlight in this paper: algorithmic and machine learning bias.
+intuitive, here we explicitly define algorithmic bias as a form of bias that
+occurs mathematical rules favors one set of attributes over others in relation
+to some target variable, like ``approve'' or ``deny'' a loan.
 
-Algorithmic bias is when a set of mathematical rules favors one set of
-attributes over others in relation to some target variable, like ``approve'' or
-``deny'' a loan. Machine learning bias is when a trained model systematically
-generates predictions that favor one group over another in relation to some
-set of attributes, e.g., education, and some target variable, e.g., ``default
-on credit''. While these definitions of bias are amoral, discrimination is in
-essence moral, occurring when an action is based on biases resulting in the
-unfair treatment of people. We define fairness as the inverse of
-discrimination, meaning that a ``fairness-aware'' model is one that produces
+Algorithmic bias in machine learning models can occur when a trained model
+systematically generates predictions that favor one group over another in
+relation to some set of attributes, e.g., education, and some target variable,
+e.g., ``default on credit''. While the above definition of bias is amoral,
+discrimination is in essence moral, occurring when an action is based on biases
+resulting in the unfair treatment of people. We define fairness as the inverse
+of discrimination, meaning that a ``fairness-aware'' model is one that produces
 non-discriminatory predictions.
@@
   \small\noindent\begin{tabularx}{\linewidth}{l X X}
-    \textbf{API Interface} & \textbf{Function} & \textbf{E\(X_a\)mples} \\
+    \textbf{API Interface} & \textbf{Function} & \textbf{Examples} \\
     \hline
@@
 The Scorer interface is concerned with measuring the degree to which data or
-predictions are discriminatory. Themis-ml implements two methods for measuring
-group-level discrimination and two methods for measuring individual-level
-discrimination.
+predictions are potentially discriminatory (PD). Themis-ml implements two
+methods for measuring group-level discrimination and two methods for measuring
+individual-level discrimination.
 
@@
 \(y^{+}\) labels and all \(X_d\) have \(y^{-}\) labels). Normalized mean
-difference scales these values to 0 and 1 based on the maximum possible
-discrimination in a dataset given the rate of positive labels
-\cite{zliobaite2015survey}.
+difference, which also takes on values between -1 and 1, scales these values
+based on the maximum possible discrimination in a dataset given the rate of
+positive label \cite{zliobaite2015survey}.
 
@@
 
-\emph{\textbf{Work in progress}. Using the German Credit dataset, we evaluate
-the efficacy of themis-ml in reducing discriminatory predictions. We do this by
-comparing the performance of various fairness-aware techniques with a standard
-``fairness- unaware'' classification system, as measured by the four
-discrimination measures described above.}
+In this section we use the German Credit dataset \cite{bache2013uci}, mean
+difference as the ``fairness'' measure, and area under the ROC curve (AUC) as
+the ``utility'' measure. The following analysis is by no means meant to be a
+comprehensive investigation of all of the possible workflows that themis-ml
+enables. Rather, it is intended to demonstrate the potential of themis-ml as a
+tool that facilites fairness-aware machine learning by enabling the user to:
+
+\begin{enumerate}
+  \item measure PD target label distributions in the training data.
+  \item measure PD predictions in a machine learning algorithm's predictions.
+  \item reduce PD predictions using fairness-aware techniques.
+  \item diagnose the fairness-utility tradeoff in a particular data context.
+\end{enumerate}
+
+The German Credit dataset classifies 1000 anonymized individuals as having
+``good'' and ``bad'' credit risks as part of a bank loan application, which we
+encode as (\(1\)) and (\(0\)) respectively to define the \emph{credit risk}
+target variable.
+
+Each individual is associated with twenty attributes such as the purpose of the
+loan, employment status, and other personal information. We begin the analysis
+by extracting three protected class attributes --- \emph{female},
+\emph{foreign worker}, and \emph{age below 25} --- and encode them as binary
+variables such that the putatively disadvantaged group is encoded as \(1\) and
+the advantaged group is encoded as \(0\) (the advantaged group would be
+\emph{male}, \emph{citizen worker}, and \emph{age above 25} respectively).
+
+Using the \textbf{Scorer} interface, we measure PD patterns with respect to
+\emph{credit risk} and each of the protected classes defined above using the
+\emph{mean difference} and \emph{normalized mean difference} metrics.
+
+\textbf{Table 3} shows the PD distribution of ``good'' and ``bad'' credit risks
+with respect to the protected attributes \emph{female}, \emph{foreign
+worker}, and \emph{age below 25}. The mean difference (md) and normalized mean
+difference (nmd) scores greater than zero suggests that the probability
+of being classified as having ``good'' risk is higher in the advantaged group
+than that of the disadvantaged group.
+
+\begin{table}
+  \caption{Potentially discriminatory target variable distribution.
+    \emph{md} = mean difference, \emph{nmd} = normalized mean difference.}
+  \renewcommand{\arraystretch}{1.75}
+  \small\noindent\begin{tabularx}{\linewidth}{l|X|X|X|X}
+    \textbf{protected class} & \textbf{md (\%)} & \textbf{md 95\% CI} &
+      \textbf{nmd (\%)} & \textbf{nmd 95\% CI}\\
+    \hline
+    \textbf{female} & 7.48 & (1.35, 13.61) & 7.73 & (1.39, 14.06) \\
+    \textbf{foreign worker} & 19.93 & (4.91, 34.94) & 63.96 & (15.76, 112.17)\\
+    \textbf{age below 25} & 14.94 & (7.76, 22.13) & 17.29 & (8.97, 25.61)\\
+  \end{tabularx}
+\end{table}
+
+\subsection{Experimental Procedure}
+
+To assess the extent to which (i) a model trained on these data mirrors these PD
+\emph{credit risk} distributions and (ii) fairness-aware techniques can reduce
+these methods, we used \emph{mean difference} to measure model fairness and
+\emph{AUC} to measure model utility. For this experiment we specify five
+conditions:
+
+\begin{itemize}
+  \item Baseline (\emph{B}): Train a model on all available input variables in
+        the German Credit dataset, including protected attributes.
+  \item Remove Protected Attribute (\emph{RPA}): Train a model on input variables
+        without protected attributes. This is the naive fairness-aware approach.
+  \item Relabel Target Variable (\emph{RTV}): Train a model using the
+        \emph{relabelling} fairness-aware method.
+  \item Counterfactually Fair Model (\emph{CFM}): Train a model using the
+        \emph{additive counterfactually fair} method.
+  \item Reject-option Classification (\emph{ROC}): Train a model using the
+        \emph{reject-option classification} method.
+\end{itemize}
+
+For each of these conditions, we train LogisticRegression, DecisionTree, and
+RandomForest model types using 10-fold cross validation, generate train and test
+predictions, and compute \emph{AUC} and \emph{mean difference} metrics for each
+train-test pair. We then compute the mean these metrics for each condition and
+model type.
+
+\subsection{Measuring and Mitigating Potentially Discriminatory Predictions}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8.5cm]{IMG/fairness_aware_comparison.png}
+\caption{
+  \textbf{Comparison of Fairness-aware Methods} using LogisticRegression,
+  DecisionTree, and RandomForest as base estimators as measured by AUC and
+  mean difference evaluated on test set predictions.}
+\label{figure-1-comparison-fa-methods}
+\end{figure}
+
+\textbf{Figure 1} suggests that in the case of LogisticRegression, the baseline
+model \emph{B} does indeed mirror the PD patterns found in the true target
+variable. Furthermore, each of the fairness-aware method appear to have
+the desired effect of reducing \emph{mean difference}, but to variance degrees
+depending on the method and protected attribute. In the \emph{female} context,
+where there appear to be the least PD (mean difference of \(7.48\%\)), the
+reductive effect of the fairness-aware methods does not appear to be as
+large as in the \emph{foreign worker} and \emph{age below 25} contexts.
+
+The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA}
+with respect to the \emph{foreign worker} protected attribute and
+LogisticRegression model illustrates the observation made by
+\cite{kusner2017counterfactual} that removing protected attributes from the
+training data does not necessarly prevent the learning algorithm address from
+mirroring PD patterns in the data.
+
+However, as the sizeable reduction in \emph{mean difference} between \emph{B}
+and \emph{RPA} with respect to the \emph{age below 25} protected attribute and
+LogisticRegression model shows, removing protected attributes can make models
+more fair while also retaining predictive power.
+
+An interesting thing to note here is that the \emph{additive counterfactually
+fair} method actually increases \emph{mean difference} for DecisionTrees and
+RandomForests across all protected attribute contexts. A possible explanation
+behind this observation is that certain assumptions made by the \emph{additive
+counterfactually fair} method are not suitable for non-linear learning
+algorithms, but this is an open question worth future inquiry.
+
+\subsection{The Fairness-utility Tradeoff}
+
+Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
+machine learning research and application, the fairness-utility tradeoff can
+help machine learning practitioners and researchers understand how effective
+particular fairness-aware methods might be in particular data contexts.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8.5cm]{IMG/fairness_utility_tradeoff.png}
+\caption{
+  \textbf{Correlation between AUC and Mean Difference} for each experimental
+  condition, across all model types (LogisticRegression, DecisionTree,
+  RandomForest). 95\% confidence intervals are provided for the pearson \(R\)
+  correlation metric.
+  }
+\label{figure-2-fairness-utility-tradeoff}
+\end{figure}
+
+\textbf{Figure 2} demonstrates an initial attempt at visualizing the the
+fairness-utility tradeoff, in this case as measured by \emph{mean difference}
+and \emph{AUC} respectively. These results suggest that the relationship between
+fairness and utility is noisy, however there does seem to be a positive
+correlation between \emph{mean difference} and \emph{AUC}, or a negative
+correlation between fairness and utility (since lower scores are better
+for \emph{mean difference} and higher scores are better for \emph{AUC}).
+
+Depending on one's use cases, analyses like this might be prove to be a useful
+guide for figuring out what kinds of methods are robust in the sense that
+one can reduce PD predictions with little to no adverse impact on predictive
+performance.
 
@@
 
-In this paper, we describe FMLIs in the classification context where we consider
-only a single binary protected class variable. More work needs to be done to
-generalize FMLIs to the multi-classification, regression, and multiple protected
-classes contexts. Additionally, little is understood about the question of the
-composability of fairness-aware methods. In other words, when different
-techniques are used together in sequence, are the resulting discrimination
-reductions additive or otherwise?
+In this paper, we describe and evaluate an FMLI in the classification context
+where we consider only a single binary protected class variable and a binary
+target variable. More work needs to be done to generalize FMLIs to the multi-
+classification, regression, and multiple protected classes contexts.
+
+Furthermore, many basic questions about model tuning, model evaluation, and
+model selection in relation to the fairness-utility tradeoff remain open. For
+instance, what might be some reasonable ways to aggregate utility and
+performance metrics in order to find the optimal set of hyperparameters?
+Additionally, little is understood about the question of the composability of
+fairness-aware methods. In other words, when different techniques are used
+together in sequence, are the resulting discrimination reductions additive or
+otherwise?
 
@@
 
-% \begin{figure}[h]
-% \centering
-% \includegraphics[width=4cm]{IMG/d4gx-2015-logo.png}
-% \caption{Conference logo for 2015 (PNG)}
-% \label{d4gx-logo-2015}
-% \end{figure}
-
-% \begin{figure*}
-%   \includegraphics[width=\textwidth]{IMG/d4gx-2015-logo.pdf}
-%   \caption{Same logo, full width.}
-% \end{figure*}
-
-% \begin{figure}[h]
-% \centering
-% \includegraphics[width=7cm]{IMG/d4gx-2015-logo.pdf}
-% \caption{Conference logo for 2015 (PDF)}
-% \label{d4gx-logo-2015}
-% \end{figure}
-
 \nocite{*}

### 2017-09-06 paper/main.tex
added paper and notebook evaluation
had to modify some of the metrics, and
also added a fairness-aware meta estimator
for convenience

@@
 
-\subsection{Experimental Procedure}
+\subsection{Measure and Mitigate Potentially Discriminatory Predictions}
+To assess the extent to which a model trained on these data mirrors these PD
+\emph{credit risk} distributions, we use the following experimental procedure:
 
-To assess the extent to which (i) a model trained on these data mirrors these PD
-\emph{credit risk} distributions and (ii) fairness-aware techniques can reduce
-these methods, we used \emph{mean difference} to measure model fairness and
-\emph{AUC} to measure model utility. For this experiment we specify five
-conditions:
-
-\begin{itemize}
-  \item Baseline (\emph{B}): Train a model on all available input variables in
-        the German Credit dataset, including protected attributes.
-  \item Remove Protected Attribute (\emph{RPA}): Train a model on input variables
-        without protected attributes. This is the naive fairness-aware approach.
-  \item Relabel Target Variable (\emph{RTV}): Train a model using the
-        \emph{relabelling} fairness-aware method.
-  \item Counterfactually Fair Model (\emph{CFM}): Train a model using the
-        \emph{additive counterfactually fair} method.
-  \item Reject-option Classification (\emph{ROC}): Train a model using the
-        \emph{reject-option classification} method.
-\end{itemize}
-
-For each of these conditions, we train LogisticRegression, DecisionTree, and
-RandomForest model types using 10-fold cross validation, generate train and test
-predictions, and compute \emph{AUC} and \emph{mean difference} metrics for each
-train-test pair. We then compute the mean these metrics for each condition and
-model type.
-
-\subsection{Measuring and Mitigating Potentially Discriminatory Predictions}
+\begin{enumerate}
+  \item \textbf{Models:} LogisticRegression, DecisionTree, RandomForest
+  \item train model using leave-one-out 10-fold cross validation
+  \item measure PD predictions in a machine learning algorithm's predictions.
+  \item reduce PD predictions using fairness-aware techniques.
+  \item diagnose the fairness-utility tradeoff in a particular data context.
+\end{enumerate}
+train a logistic regression model using
+leave-one-out 10-fold cross validation, generate predictions on the training and
+test folds, and compute \emph{AUC} and \emph{mean difference} for each
+train-test partition. We then compute the mean of these performance metrics
+over a 1-dimensional grid of L2 regularization hyperparemeter settings.
 
@@
 \caption{
-  \textbf{Comparison of Fairness-aware Methods} using LogisticRegression,
-  DecisionTree, and RandomForest as base estimators as measured by AUC and
-  mean difference evaluated on test set predictions.}
+  \textbf{Comparison of Fairness-aware Methods}
+  % are shown for each of the three protected class contexts: \emph{female},
+  % \emph{foreign worker}, and \emph{age below 25}. Left panels show training and
+  % test \emph{AUC} scores (higher is better) of an L2-regularized Logistic
+  % Regression model. The \(C\) parameter is the inverse regularization strength,
+  % meaning that smaller values increase the regularization penalty. Model
+  % complexity increases as \(C\) increases, leading to the overfitting regime at
+  % \(C > 10^{-1}\). Right panels show training and test \emph{mean difference}
+  % scores (lower is better).
+  }
 \label{figure-1-comparison-fa-methods}
 \end{figure}
-
-\textbf{Figure 1} suggests that in the case of LogisticRegression, the baseline
-model \emph{B} does indeed mirror the PD patterns found in the true target
-variable. Furthermore, each of the fairness-aware method appear to have
-the desired effect of reducing \emph{mean difference}, but to variance degrees
-depending on the method and protected attribute. In the \emph{female} context,
-where there appear to be the least PD (mean difference of \(7.48\%\)), the
-reductive effect of the fairness-aware methods does not appear to be as
-large as in the \emph{foreign worker} and \emph{age below 25} contexts.
-
-The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA}
-with respect to the \emph{foreign worker} protected attribute and
-LogisticRegression model illustrates the observation made by
-\cite{kusner2017counterfactual} that removing protected attributes from the
-training data does not necessarly prevent the learning algorithm address from
-mirroring PD patterns in the data.
-
-However, as the sizeable reduction in \emph{mean difference} between \emph{B}
-and \emph{RPA} with respect to the \emph{age below 25} protected attribute and
-LogisticRegression model shows, removing protected attributes can make models
-more fair while also retaining predictive power.
-
-An interesting thing to note here is that the \emph{additive counterfactually
-fair} method actually increases \emph{mean difference} for DecisionTrees and
-RandomForests across all protected attribute contexts. A possible explanation
-behind this observation is that certain assumptions made by the \emph{additive
-counterfactually fair} method are not suitable for non-linear learning
-algorithms, but this is an open question worth future inquiry.
-
-\subsection{The Fairness-utility Tradeoff}
-
-Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
-machine learning research and application, the fairness-utility tradeoff can
-help machine learning practitioners and researchers understand how effective
-particular fairness-aware methods might be in particular data contexts.
 
@@
 \caption{
-  \textbf{Correlation between AUC and Mean Difference} for each experimental
-  condition, across all model types (LogisticRegression, DecisionTree,
-  RandomForest). 95\% confidence intervals are provided for the pearson \(R\)
-  correlation metric.
+  \textbf{Diagnosing the Fairness-utility Tradeoff}
+  % are shown for each of the three protected class contexts: \emph{female},
+  % \emph{foreign worker}, and \emph{age below 25}. Left panels show training and
+  % test \emph{AUC} scores (higher is better) of an L2-regularized Logistic
+  % Regression model. The \(C\) parameter is the inverse regularization strength,
+  % meaning that smaller values increase the regularization penalty. Model
+  % complexity increases as \(C\) increases, leading to the overfitting regime at
+  % \(C > 10^{-1}\). Right panels show training and test \emph{mean difference}
+  % scores (lower is better).
   }
@@
 
-\textbf{Figure 2} demonstrates an initial attempt at visualizing the the
-fairness-utility tradeoff, in this case as measured by \emph{mean difference}
-and \emph{AUC} respectively. These results suggest that the relationship between
-fairness and utility is noisy, however there does seem to be a positive
-correlation between \emph{mean difference} and \emph{AUC}, or a negative
-correlation between fairness and utility (since lower scores are better
-for \emph{mean difference} and higher scores are better for \emph{AUC}).
+% \begin{figure}
+%   \centering
+%   \begin{subfigure}[t]{1in}
+%     \centering
+%     \includegraphics[width=6cm]{IMG/logistic_regression_validation_curve.png}
+%     \caption{Arabic numerals}\label{fig:1a}
+%   \end{subfigure}
+%   \quad
+%   \begin{subfigure}[t]{1in}
+%     \centering
+%     \includegraphics[width=3cm]{IMG/logistic_regression_fairness_utility_tradeoff.png}
+%     \caption{Arabic numerals}\label{fig:1b}
+%   \end{subfigure}
+%   \caption{Capital Roman numerals}\label{fig:1}
+% \end{figure}
 
-Depending on one's use cases, analyses like this might be prove to be a useful
-guide for figuring out what kinds of methods are robust in the sense that
-one can reduce PD predictions with little to no adverse impact on predictive
-performance.
+In the machine learning field we often talk about the bias-variance tradeoff,
+which refers to the observation that simple models exhibit high bias and low
+variance, i.e. they ``underfit'' the data by consistently performing poorly on
+both the test and training set across many instantiations of the model.
+
+However, as model complexity increases (for example, by introducing first order
+polynomial terms into space), models begin to exhibit low bias and high
+variance. This means that they ``overfit'' the data by performing well on the
+training set but poorly on the test set.
+
+In practice, we often seek to find the optimal model specification that
+maximizes/minimizes test set performance/error but also minimizes the gap
+between test set and training set performance/error.
+
+A concept that is related but distinct from the bias-variance tradeoff is the
+fairness-utility tradeoff, which is the observation that fair models tend to
+make less accurate predictions \cite{kamiran2012data, dwork2012fairness}. Just
+as the bias-variance tradeoff has become a useful diagnostic tool to guide
+machine learning research and application, the fairness-utility tradeoff can
+help .
 
@@
 
-In this paper, we describe and evaluate an FMLI in the classification context
-where we consider only a single binary protected class variable and a binary
-target variable. More work needs to be done to generalize FMLIs to the multi-
-classification, regression, and multiple protected classes contexts.
-
-Furthermore, many basic questions about model tuning, model evaluation, and
-model selection in relation to the fairness-utility tradeoff remain open. For
-instance, what might be some reasonable ways to aggregate utility and
-performance metrics in order to find the optimal set of hyperparameters?
-Additionally, little is understood about the question of the composability of
-fairness-aware methods. In other words, when different techniques are used
-together in sequence, are the resulting discrimination reductions additive or
-otherwise?
+In this paper, we describe FMLIs in the classification context where we consider
+only a single binary protected class variable. More work needs to be done to
+generalize FMLIs to the multi-classification, regression, and multiple protected
+classes contexts. Additionally, little is understood about the question of the
+composability of fairness-aware methods. In other words, when different
+techniques are used together in sequence, are the resulting discrimination
+reductions additive or otherwise?
 

### 2017-09-06 tests/test_meta_estimators.py

@@ -1,0 +1,80 @@
+"""Unit tests for themis_ml meta estimators."""
+
+import pytest
+
+from sklearn.linear_model import LogisticRegression
+
+from themis_ml.meta_estimators import FairnessAwareMetaEstimator
+from themis_ml.linear_model import counterfactually_fair_models
+from themis_ml.preprocessing import relabelling
+from themis_ml.postprocessing import reject_option_classification
+
+from conftest import create_linear_X, create_y, create_s
+
+
+def test_fairness_aware_meta_estimator():
+    X = create_linear_X()
+    y = create_y()
+    s = create_s()
+
+    # use sklearn estimator
+    lr = FairnessAwareMetaEstimator(LogisticRegression())
+    lr.fit(X, y)
+    lr.predict(X)
+    lr.predict_proba(X)
+
+    # use themis_ml LinearACFClassifier estimator
+    linear_acf = FairnessAwareMetaEstimator(
+        counterfactually_fair_models.LinearACFClassifier(
+            binary_residual_type="absolute"))
+    linear_acf.fit(X, y, s)
+    linear_acf.predict(X, s)
+    linear_acf.predict_proba(X, s)
+
+    # when fit/predict methods need s, raise ValueError
+    with pytest.raises(ValueError):
+        linear_acf.fit(X, y, s=None)
+    with pytest.raises(ValueError):
+        linear_acf.predict(X, s=None)
+    with pytest.raises(ValueError):
+        linear_acf.predict_proba(X, s=None)
+
+    # use themis_ml relabeller preprocessor
+    relabel_clf = FairnessAwareMetaEstimator(
+        estimator=LogisticRegression(), relabeller=relabelling.Relabeller())
+    relabel_clf.fit(X, y, s)
+    relabel_clf.predict(X)
+    relabel_clf.predict_proba(X)
+
+    # when estimator method does not need `s` on predict, raise ValueError if
+    # it is provided
+    with pytest.raises(ValueError):
+        relabel_clf.predict(X, s)
+        relabel_clf.predict_proba(X, s)
+
+    # use themis_ml RejectOption Classifier
+    single_reject_option_clf = FairnessAwareMetaEstimator(
+        reject_option_classification.SingleROClassifier())
+    single_reject_option_clf.fit(X, y)
+    single_reject_option_clf.predict(X, s)
+    single_reject_option_clf.predict_proba(X, s)
+
+    with pytest.raises(ValueError):
[…]

### 2017-09-06 tests/test_metrics.py

@@ -5,44 +5,49 @@ import pytest
 from themis_ml import metrics
 
 
+def _get_point_est(md_tuple):
+    """Get point estimate in mean_difference/normalized_mean_difference."""
+    return md_tuple[0]
+
+
 def test_mean_difference_full_discrimination():
     """Binary case: disadvantaged group is fully discriminated against."""
     y = [1, 1, 1, 1, 0, 0, 0, 0]
     s = [0, 0, 0, 0, 1, 1, 1, 1]
-    assert metrics.mean_difference(y, s) == 1
-    assert metrics.normalized_mean_difference(y, s) == 1
+    assert _get_point_est(metrics.mean_difference(y, s)) == 1
+    assert _get_point_est(metrics.normalized_mean_difference(y, s)) == 1
 
 
 def test_mean_difference_partial_discrimination():
     """Binary case: disadvantaged group is partially discriminated against."""
     y = [1, 1, 1, 1, 0, 0, 1, 1]
     s = [0, 0, 0, 0, 1, 1, 1, 1]
-    assert metrics.mean_difference(y, s) == 0.5
-    assert metrics.normalized_mean_difference(y, s) == 1
+    assert _get_point_est(metrics.mean_difference(y, s)) == 0.5
+    assert _get_point_est(metrics.normalized_mean_difference(y, s)) == 1
 
 
 def test_mean_difference_no_discrimination():
     """Binary case: proportion in each group with +ve outcome are equal."""
     y = [0, 0, 1, 1, 0, 0, 1, 1]
     s = [0, 0, 0, 0, 1, 1, 1, 1]
-    assert metrics.mean_difference(y, s) == 0
-    assert metrics.normalized_mean_difference(y, s) == 0
+    assert _get_point_est(metrics.mean_difference(y, s)) == 0
+    assert _get_point_est(metrics.normalized_mean_difference(y, s)) == 0
 
 
 def test_mean_difference_partial_reverse_discrimination():
     """Binary case: advantaged group is partially discriminated against."""
     y = [1, 1, 0, 0, 1, 1, 1, 1]
     s = [0, 0, 0, 0, 1, 1, 1, 1]
-    assert metrics.mean_difference(y, s) == -0.5
-    assert metrics.normalized_mean_difference(y, s) == -1
+    assert _get_point_est(metrics.mean_difference(y, s)) == -0.5
+    assert _get_point_est(metrics.normalized_mean_difference(y, s)) == -1
 
 
[…]

### 2017-09-06 themis_ml/checks.py

@@ -21,3 +21,22 @@ def is_binary(x):
 
 def is_continuous(x):
     return not is_binary(x) and x.dtype in CONTINUOUS_DTYPES
+
+
+def s_is_needed_on_fit(estimator, s):
+    if getattr(estimator, "S_ON_FIT", False):
+        if s is None:
+            raise ValueError(
+                "Provide `s` arg when calling %s's `fit`" % estimator)
+        return True
+    return False
+
+
+def s_is_needed_on_predict(estimator, s):
+    if getattr(estimator, "S_ON_PREDICT", False):
+        if s is None:
+            raise ValueError(
+                "Provide `s` arg when calling %s's `predict`" %
+                estimator)
+        return True
+    return False

### 2017-09-06 themis_ml/linear_model/counterfactually_fair_models.py

@@ -59,6 +59,8 @@ class _BinaryResidualTypes(Enum):
 class LinearACFClassifier(BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
 
     VALID_BINARY_RESIDUAL_TYPES = [e.name for e in _BinaryResidualTypes]
+    S_ON_FIT = True
+    S_ON_PREDICT = True
 
     def __init__(self, target_estimator=LogisticRegression(),
                  continuous_estimator=LinearRegression(),
@@ -68,8 +70,8 @@ class LinearACFClassifier(BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
 
         :param BaseEstimator target_estimator: A classifier for learning a
             function that maps to input residuals and target variable.
-        :param BaseEstimat0r linear_estimator: A regressor for computing the
-            residuals for continuous X inputs.
+        :param BaseEstimator continuous_estimator: A regressor for computing
+            the residuals for continuous X inputs.
         :param BaseEstimator binary_estimator: A classifier estimator for
             computing the residuals for binary X inputs.
         :param str binary_residual_type: The type of residual to use for binary
@@ -83,7 +85,7 @@ class LinearACFClassifier(BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
         self.target_estimator = target_estimator
         self.continuous_estimator = continuous_estimator
         self.binary_estimator = binary_estimator
-        self.binary_residual_type = _BinaryResidualTypes[binary_residual_type]
+        self.binary_residual_type = binary_residual_type
 
     def fit(self, X, y, s):
         """Fit model."""
@@ -117,7 +119,7 @@ class LinearACFClassifier(BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
                 estimator = clone(self.binary_estimator)
                 compute_residual_func = partial(
                     _compute_binary_residuals,
-                    residual_type=self.binary_residual_type)
+                    residual_type=self._binary_residual_type)
             else:
                 raise ValueError(
                     "index %s is not in continuous_index_ or binary_index_")
@@ -173,3 +175,7 @@ class LinearACFClassifier(BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
         self._check_fitted(X)
         predict_residuals = self._compute_residuals_on_predict(X, s)
         return self.target_estimator_.predict_proba(predict_residuals)
+
+    @property
+    def _binary_residual_type(self):
+        return _BinaryResidualTypes[self.binary_residual_type]

### 2017-09-06 themis_ml/meta_estimators.py

@@ -1,0 +1,72 @@
+"""Module for Fairness-aware base estimators."""
+
+import numpy as np
+
+from sklearn.base import (
+    BaseEstimator, ClassifierMixin, MetaEstimatorMixin, clone)
+from sklearn.utils.validation import check_array, check_X_y, check_is_fitted
+from checks import check_binary, s_is_needed_on_fit, s_is_needed_on_predict
+
+
+class FairnessAwareMetaEstimator(
+        BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
+
+    def __init__(self, estimator, relabeller=None):
+        """Initialize metaestimator for composing fairness-aware methods.
+
+        :param Estimator estimator:
+        :param Transformer|None relabeller:
+        """
+        self.relabeller = relabeller
+        self.estimator = estimator
+
+    def fit(self, X, y, s=None):
+        X, y = check_X_y(X, y)
+        y = check_binary(y)
+        self.relabeller_ = None
+        self.estimator_ = clone(self.estimator)
+        # fit_transform y labels using estimator
+        if self.relabeller is not None:
+            self.relabeller_ = clone(self.relabeller)
+            y = self.relabeller_.fit_transform(X, y, s=s)
+        # fit estimator
+        if s_is_needed_on_fit(self.estimator_, s):
+            s = check_binary(np.array(s).astype(int))
+            self.estimator_.fit(X, y, s)
+        else:
+            # since relabeller by definition needs s, this checks whether
+            # relabeller is None and the `s` array is provided.
+            if self.relabeller_ is None and s is not None:
+                raise ValueError(
+                    "`s` arg provided but %s fit doesn't accept `s`" %
+                    self.estimator_)
+            self.estimator_.fit(X, y)
+
+    def predict(self, X, s=None):
+        check_is_fitted(self, ["estimator_", "relabeller_"])
+        X = check_array(X)
+        if s_is_needed_on_predict(self.estimator_, s):
+            s = check_binary(np.array(s).astype(int))
+            return self.estimator_.predict(X, s)
+        else:
+            if s is not None:
+                raise ValueError(
+                    "`s` arg provided but %s predict doesn't accept `s`" %
+                    self.estimator_)
+            return self.estimator_.predict(X)
+
+    def predict_proba(self, X, s=None):
+        if not hasattr(self.estimator_, "predict_proba"):
+            raise AttributeError(
+                "%s has no method `predict_proba`" % self.estimator_)
[…]

### 2017-09-06 themis_ml/metrics.py

@@ -3,6 +3,33 @@
 import numpy as np
 
 from .checks import check_binary
+from math import sqrt
+from scipy.stats import t
+
+
+def mean_differences_ci(y, s, ci=0.975):
+    """Calculate the mean difference and confidence interval.
+
+    :param array-like y: shape (n, ) containing binary target variable, where
+        1 is the desireable outcome and 0 is the undesireable outcome.
+    :param array-like s: shape (n, ) containing binary protected class
+        variable where 0 is the advantaged groupd and 1 is the disadvantaged
+        group.
+    :param float ci: % confidence interval to compute. Default: 97.5% to
+        compute 95% two-sided t-statistic associated with degrees of freedom.
+    :returns: mean difference between advantaged group and disadvantaged group
+        with lower and upper bound confidence interval estimates.
+    :rtype: tuple
+    """
+    n0 = (s == 0).sum().astype(float)
+    n1 = (s == 1).sum().astype(float)
+    df = n0 + n1 - 2
+    std0 = y[s == 0].std()
+    std1 = y[s == 1].std()
+    std_n0n1 = sqrt(((n1 - 1)*(std1)**2 + (n0 - 1)*(std0)**2) / df)
+    mean_diff = y[s == 0].mean() - y[s == 1].mean()
+    margin_error = t.ppf(ci, df) * std_n0n1 * sqrt(1/n0 + 1 / float(n1))
+    return mean_diff, mean_diff - margin_error, mean_diff + margin_error
 
 
 def _mean_difference(y, s):
@@ -30,9 +57,9 @@ def mean_difference(y, s):
     Zliobaite, I. (2015). A survey on measuring indirect discrimination in
     machine learning. arXiv preprint arXiv:1511.00148.
 
-    :param array-like y: shape (n, ) containing binary target variable, where
+    :param numpy.array y: shape (n, ) containing binary target variable, where
         1 is the desireable outcome and 0 is the undesireable outcome.
-    :param array-like s: shape (n, ) containing binary protected class
+    :param numpy.array s: shape (n, ) containing binary protected class
         variable where 0 is the advantaged groupd and 1 is the disadvantaged
         group.
     :returns: mean difference between advantaged group and disadvantaged group.
@@ -40,10 +67,10 @@ def mean_difference(y, s):
     """
     y = check_binary(np.array(y).astype(int))
     s = check_binary(np.array(s).astype(int))
-    return _mean_difference(y, s)
[…]

### 2017-09-06 themis_ml/postprocessing/reject_option_classification.py

@@ -19,6 +19,9 @@ DEFAULT_ENSEMBLE_ESTIMATORS = [
 class SingleROClassifier(
         BaseEstimator, ClassifierMixin, MetaEstimatorMixin):
 
+    S_ON_FIT = False
+    S_ON_PREDICT = True
+
     def __init__(self, estimator=LogisticRegression(), theta=0.1, demote=True):
         """Initialize Single Reject-Option Classifier.
 

### 2017-11-29 paper/journal_tech_human_services/main_interact.tex
build for journal of tech in hum serv

\documentclass{interact}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{apacite}
\usepackage{natbib}
\usepackage[labelfont=bf]{caption}
\bibliographystyle{apacite}
\captionsetup{labelfont = bf}
\setlength{\parskip}{1em}

\begin{document}

% define code block framing style
\BeforeBeginEnvironment{minted}{
  \begin{mdframed}[skipabove=20pt,
                   skipbelow=10pt,
                   linewidth=0pt,
                   innerleftmargin=10pt,
                   innertopmargin=0pt,
                   innerbottommargin=0pt]}
\AfterEndEnvironment{minted}{\end{mdframed}}

\title{
  Themis-ml: A Fairness-aware Machine Learning Interface for \\
  End-to-end Discrimination Discovery and Mitigation
}

% \author{
%   \name{Niels Bantilan\textsuperscript{a}\thanks{Contact: Niels Bantilan. Email: niels.bantilan@gmail.com}}
%   \affil{\textsuperscript{a}Arena.io, New York, NY}
% }

% For the draft version
\author{
  \name{Anonymous\textsuperscript{a}\thanks{Contact: anonymous. Email: anonymous}}
  \affil{\textsuperscript{a}no affiliation}
}

\maketitle

\begin{abstract}
As more industries integrate machine learning into socially sensitive decision
processes like hiring, loan-approval, and parole-granting, we are at risk of
perpetuating historical and contemporary socioeconomic disparities. This is a
critical problem because on the one hand, organizations who use but do not
understand the discriminatory potential of such systems will facilitate the
widening of social disparities under the assumption that algorithms are
categorically objective. On the other hand, the responsible use of machine
learning can help us measure, understand, and mitigate the implicit historical
biases in socially sensitive data by expressing implicit decision-making mental
models in terms of explicit statistical models. In this paper we specify,
implement, and evaluate a ``fairness-aware'' machine learning interface called
themis-ml, which is intended for use by individual data scientists and
engineers, academic research teams, or larger product teams who use machine
learning in production systems.
\end{abstract}

\begin{keywords}
Fairness-aware machine learning; discrimination discovery; statistics;
social bias; application programming interface
\end{keywords}

\section{Introduction} In recent years, the transformative potential of machine
learning (ML) in many industries has propelled ML into the forefront of
mainstream media. From improving products and services to optimizing logistics
and operations, ML and artificial intelligence more broadly offer a wide range
of tools for organizations to enhance their internal and external capabilities.

As with any tool, we can use ML to engender great social benefit, but as
\citet{o2017weapons} emphasizes, we can also misuse it to bring about devastating
harm. In this paper, we focus on ML systems in the context of Decision Support
Systems (DSS), which are software systems that are intended to assist humans in
various decision-making contexts \cite{yoshimura2006decision,
montgomery2000evaluation, barnett1987dxplain, mysiak2005towards}. The misuse of
ML in these types of systems could potentially precipitate a widespread adverse
impact on society by introducing insidious feedback loops between biased
historical data and current decision-making \cite{o2017weapons}.

Researchers have developed many discrimination discovery and fairness-aware ML
methods (\citealp{kusner2017counterfactual, kamiran2012data, kamishima2012fairness,
kamiran2012decision, zemel2013learning, zafar2017fairness, dwork2012fairness,
zliobaite2015survey}), so we build on work done by others and seek to leverage
these techniques in the context of research- and product-based machine learning
applications.

Our contributions in this paper are three-fold. First, we propose an application
programming interface (API) for ``Fairness-aware Machine Learning Interfaces''
(FMLI) in the context of a simple binary classifier. Second, we introduce
themis-ml, an FMLI-compliant library, and apply it to a hypothetical
loan-granting DSS using the German Credit Dataset \cite{bache2013uci}. Finally, we
evaluate the efficacy of themis-ml as a tool for measuring potential
discrimination (PD) in both training data and ML predictions as well as
mitigating PD using fairness-aware methods. Our hope is that themis-ml serves as
a reference implementation that others might use and extend for their own
purposes.

\section{Bias and Discrimination}
Colloquially, bias is simply a preference for or against something, e.g.
preferring vanilla over chocolate ice cream. While this definition is
intuitive, here we explicitly define algorithmic bias as a form of bias that
occurs when mathematical rules favor one set of attributes over others in
relation to some target variable, like ``approving'' or ``denying'' a loan.

Algorithmic bias in machine learning models can occur when a trained model
generates predictions that favor one group over another in relation to some set
of attributes, e.g. education, and some target variable, e.g. ``default on
credit''. While the definition above of bias is amoral, discrimination is in
essence moral, occurring when an action is based on biases resulting in the
systematic disenfranchisement of people based on arbitrary group membership. We
define fairness as the inverse of discrimination, meaning that a ``fairness-
aware'' model is one that produces non-discriminatory predictions.

Bias can lead to either direct (intended/explicit) or indirect
(unintended/implicit) discrimination, and the predominant legal concepts used to
determine these two types are known as disparate treatment and disparate impact,
respectively \cite{barocas2016big}. As \cite{kusner2017counterfactual,
kamiran2012data} suggest, we can address disparate treatment in ML models by
simply removing all variables that are highly correlated to the protected class
of interest, in addition to the protected class itself, from the training data.
However, as \cite{kusner2017counterfactual} points out, doing so does not
necessarily mitigate discriminatory predictions and may actually introduce
unfairness into an otherwise fair system. In contrast, addressing disparate
impact is more complex because it depends on historical processes that generated
the training data, non-linear relationships between the features and protected
class, and whether we are interested in measuring individual- or group-level
discrimination \cite{dwork2012fairness}.

\section{A Fairness-aware Machine Learning Interface}

So how does one measure
disparate impact and individual-/group-level discrimination in an ML-driven
product? In this section, we describe the main components of a simple
classification system, enumerate a few of the use cases that a research or
product team might have for using an FMLI, and propose an API that fulfills
these use cases.

A simple classification ML pipeline consists of five steps: data ingestion, data
preprocessing, model training, model evaluation, and prediction generation on
new examples. Data ingestion is outside the scope of this paper because it is a
highly variable process that depends on the application, often involves
considerable engineering effort, and potentially requires external stakeholder
buy-in.

Table 1 outlines a simple classification system in terms of the core interfaces
in scikit-learn (sklearn), which is a machine learning library in the Python
programming language \cite{scikit-learn}, and table 2 delineates some of the
use cases that research or product teams might have to justify the use of an
FMLI.

\begin{table}
  \caption{A Simple Classification Pipeline}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l X X}
    \textbf{API Interface} & \textbf{Function} & \textbf{Examples} \\
    \hline
    \textbf{Transformer} &
      \emph{Preprocess} raw data for model training. &
      mean-unit variance scaling, min-max scaling \\
    \textbf{Estimator} &
      \emph{Train} models to perform a classification task. &
      logistic regression, random forest \\
    \textbf{Scorer} &
      \emph{Evaluate} performance of different models. &
      accuracy, f1-score, area under the curve \\
    \textbf{Predictor} &
      \emph{Predict} outcomes for new data. &
      single-classifier prediction, ensemble prediction
  \end{tabularx}
\end{table}

\section{FMLI Specification}

Here we propose a high-level specification of themis-ml, an open source FMLI
named after the ancient Greek titaness of justice (the library can be found on
% \href{https://github.com/cosmicBboy/themis-ml}{\underline{github}}.)
\underline{github} [\emph{link removed to maintain anonymity}].)
We adopt sklearn's principles of consistency, inspection, non- proliferation of
classes, composition, and sensible defaults \cite{buitinck2013api}, and extend
them with the following FMLI-specific principles:

\begin{hangparas}{.15in}{1}
  \textbf{Model flexibility.} Focus on fairness-aware methods that are applicable
    to a variety of model types because users might have no control or full
    control over the specific model training implementation.
\end{hangparas}
\begin{hangparas}{.15in}{1}
  \textbf{Fairness as performance.} Provide estimators and scoring metrics that
    explicitly encode a notion of both model accuracy and fairness so that models
    can optimize for both.
\end{hangparas}
\begin{hangparas}{.15in}{1}
\textbf{Transparency of fairness-utility tradeoff.} Fair models often make less
  accurate predictions [8, 13], which is an important factor when assessing
  their business impact.
\end{hangparas}

\begin{table}
  \caption{FMLI Use Cases}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{X X}
    \textbf{Use Case} & \textbf{Rationale} \\
    \hline
    Detect and reduce discrimination in a production machine learning pipeline.
      & Fairness-aware modeling aligns with team/company values,
        provides protection from legal liability. \\
    Measure individual-/group-level discrimination in data with respect to a
    protected class and outcome of interest.
      & Need to assess the potential bias resulting from training models on
        data. \\
    Preprocess raw data or post-process model predictions in a way that reduces
    discriminatory predictions generated by models.
      & Unable to change the underlying implementation of the model training
        process. \\
    Explicitly learn model parameters that produce fair predictions for a variety
    of model types.
      & Need for flexibility when experimenting with or deploying different model
        types. \\
    Evaluate the degree to which fairness-aware methods reduce discrimination
    and assess the fairness-utility tradeoff.
      & Need for assessing the business consequences or other implications of
        deploying a fairness-aware model.
  \end{tabularx}
\end{table}

\subsection{Preliminaries} In the following subsections we describe specific
methods from the ML fairness literature that map onto each of the sklearn
interfaces. Note that we only provide a high level summary of each method,
citing the original sources for more implementation details. The following
descriptions make two assumptions: (i) the positive target label \(y^{+}\)
refers to a desirable outcome, e.g. ``approve loan'', and vice versa for the
negative target label \(y^{-}\), and (ii) the protected class is a binary
variable defined as \(s \in \{d, a\}\), where \(X_d\)  are members of the
disadvantaged group and \(X_a\) are members of the advantaged group.

Following these conventions, we define \(X_{d, y^{+}}\) and \(X_{d, y^{-}}\)
as the set of observations of the disadvantaged group that are positively
labelled and negatively labelled, respectively. Similarly, \(X_{a, y^{+}}\), and
\(X_{a, y^{-}}\)  are observations of the advantaged group that are positively
and negatively labelled, respectively.

\subsection{Transformer} The main idea behind fairness-aware preprocessing is to
take a dataset \(D\) consisting of a feature set \(X_{train}\), target labels
\(y_{train}\), and protected class \(s_{train}\) to output a modified dataset.

\emph{Relabelling}, also called \emph{Massaging}, modifies \(y_{train}\) by relabelling
the target variables in such a way that ``promotes'' members of the
disadvantaged protected class (e.g. ``immigrant'') and ``demotes'' members of
the advantaged class (e.g. ``citizen'') \cite{kamiran2012data}. A ranker \(R\)
(e.g. logistic regression) is trained on \(D\), and ranks are generated for
all observations. Some of the top-ranked observations \(X_{d, y^{-}}\)  are
``promoted'' to \(X_{d, y^{+}}\) and some of the bottom-ranked observations
\(X_{a, y^{+}}\)  are ``demoted'' to \(X_{a, y^{-}}\) such that the proportion
of \(y^{+}\) are equal in both \(X_d\) and \(X_a\). Two caveats of this method
are that it is intrusive because it directly manipulates \(y\), and that it
narrowly defines fairness as the uniform distribution of benefits between
\(X_a\) and \(X_d\).

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import Relabeller
  from sklearn.linear_model import LogisticRegression

  # use logistic regression as the ranking algorithm
  massager = Relabeller(ranker=LogisticRegression)

  # obtain a new set of labels
  new_y = massager.fit_transform(X, y, s)

  # train any model on new y labels
  lr = LogisticRegression()
  lr.fit(X, new_y)
\end{minted}

\emph{Reweighting} takes a dataset D and assigns a weight to each observation
using conditional probabilities based on \(y\) and \(s\) \cite{kamiran2012data}.
In brief, large weights are assigned to \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\) ,
while small weights are assigned to \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\). The
weights are then used as input to model types that support weighted sample
observations --- which actually points to the main limitation of this method,
since not all classifiers can incorporate observation weights during the
learning process.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import Reweight
  from sklearn.linear_model import LogisticRegression

  reweigher = Reweight()

  # obtain fairness-aware weights for each observation
  reweigher.fit(y, s)
  fair_weights = reweigher.transform(y, s)

  # train a LogisticRegression model with sample weights
  lr = LogisticRegression()
  lr.fit(X, y, weights=fair_weights)
\end{minted}

\emph{Sampling} is composed of two methods: the first involves uniformly
sampling n observations from each group, where n is the expected size of that
group assuming a uniform distribution. The second is to preferentially sample
observations using a ranker \(R\), similar to the \emph{Relabelling} method. The
procedure is to duplicate the top-ranked \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\)
while removing top-ranked \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\)
\cite{kamiran2012data}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.preprocess import (
      UniformSample, PreferentialSample)
  from sklearn.linear_model import LogisticRegression

  # use logistic regression as the ranking algorithm
  uniform_sampler = UniformSample()
  preferential_sampler = PreferentialSample(
      ranker=LogisticRegression)

  # obtain a new dataset with uniform sampling
  uniform_sampler.fit(y_train, s_train)
  X, y, s = uniform_sampler.transform(X, y, s)

  # obtain a new dataset with preferential sampling
  preferential_sampler.fit(y_train, s_train)
  X, y, s = preferential_sampler.transform(X, y, s)
\end{minted}

\subsection{Estimator}

Themis-ml implements two methods for training fairness-aware models: the
prejudice remover regularizer (PRR), and the additive counterfactually fair
(ACF) model.

\cite{kamishima2012fairness} proposes PRR as an optimization technique that
extends the standard L1/L2-norm regularization method \cite{ng2004feature,
ribeiro2016should} by adding a prejudice index term to the objective function.
This term is equivalent to normalized mutual information, which measures the
degree to which predictions \(y\) and \(s\) are dependent on each other. With
values ranging from 0 to 1, 0 means that \(y\) and \(s\) are independent, and a
value of 1 means that they are dependent. The goal of the objective function is
to find model parameters that minimize the difference between the true label
\(y\) and the predicted label \(\hat{y}\) in addition to the degree to which
\(y\) depends on s.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.linear_model import LogisticRegressionPRR

  # use L2-norm regularization and prejudice index as
  # the discrimination penalizer
  lr_prr = LogisticRegressionPRR(
      penalty="L2", discrimination_penalty="PI")

  # fit the models
  lr_prr.fit(X, y, s)
\end{minted}

ACF is a method described by \cite{kusner2017counterfactual} within the
framework of counterfactual fairness. The main idea is to train linear models to
predict each feature using the protected class attribute(s) as input. We can
then compute the residuals \(\epsilon_{ij}\) between the predicted feature
values and true feature values for each observation i and each feature j. The
final model is then trained on \(\epsilon_{ij}\) as features to predict y.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.linear_model import LinearACFClassifier

  # by default, LinearACFClassifier uses linear
  # regression as the continuous feature estimator
  # and logistic regression as the binary feature
  # estimator and target variable classifier
  linear_acf = LinearACFClassifier()

  # fit the models
  linear_acf.fit(X_train, y_train, s_train)
\end{minted}

\subsection{Predictor}

Themis-ml draws on two methods to make model type-agnostic predictions:
\emph{Reject Option Classification} (ROC) and \emph{Discrimination Aware
Ensemble Classification} (DAEC) \cite{kamiran2012decision}. Unlike the
Transformer and Estimator methods outlined above, ROC and DAEC do not modify the
training data or the training process. Rather, they postprocess predictions in a
way that reduces potentially discriminatory (PD) predictions.

\cite{kamiran2012decision} describes two ways of implementing ROC, starting with
ROC in a single classifier setting. ROC works by training an initial classifier
on \(D\), generating predicted probabilities on the test set, and then computing the
proximity of each prediction to the decision boundary learned by the classifier.
Within this boundary defined by the critical region threshold \(\theta\), where
0.5 < \(\theta\) < 1, \(X_d\) are assigned as \(y^{+}\) and \(X_a\) are assigned
as \(y^{-}\). ROC in the multiple classifier setting is similar to the single
classifier setting, except that predicted probabilities are defined as the
weighted average of probabilities generated by each classifier.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.postprocessing import (
      SingleROClassifier, MultiROClassifier)
  from sklearn.linear_model import LogisticRegression
  from sklearn.tree import DecisionTreeClassifier

  # use logistic regression for single classifier setting
  single_roc = SingleROClassifier(
      estimator=LogisticRegression())

  # use logistic regression and decision trees for
  # multiple classifier setting
  multi_roc = MultiROClassifier(
      estimators=[LogisticRegression(),
                  DecisionTreeClassifier()])

  # fit the models and generate predictions
  single_roc.fit(X, y, s)
  multi_roc.fit(X, y, s)
  single_roc.predict(X, s)
  multi_roc.predict(X, s)
\end{minted}

The main limitation of ROC is that model types must be able to produce predicted
probabilities. DAEC gets around this problem by training an ensemble of
classifiers and, through a similar relabelling rule as ROC, re-assigns any
prediction where classifiers disagree on the predicted label. As
\cite{kamiran2012decision} notes, in general, the larger the disagreement
between classifiers, the larger the reduction in discrimination.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.postprocessing import DAEnsembleClassifier
  from sklearn.linear_model import LogisticRegression
  from sklearn.tree import DecisionTreeClassifier

  # use logistic regression and decision trees
  dae_clf = DAEnsembleClassifier(
      estimators=[LogisticRegression(),
                  DecisionTreeClassifier()])

  # fit the models and generate predictions
  dae_clf.fit(X, y, s)
  dae_clf.predict(X, s)
\end{minted}

\subsection{Scorer}

The Scorer interface is concerned with measuring the degree to which data or
predictions are PD. Themis-ml implements two methods for measuring group-level
discrimination and two methods for measuring individual-level discrimination.

In the context of measuring group-level discrimination,
\cite{zliobaite2015survey} describes \emph{mean difference} and \emph{normalized
mean difference}. \emph{Mean difference} measures the difference between \(p(a
\cup y^{+})\) and \(p(d \cup y^{+})\). Values range from -1 to 1, where -1 is
the reverse-discrimination case (all \(X_a\) have \(y^{-}\) labels and all
\(X_d\) have \(y^{+}\) labels) and 1 is the fully discriminatory case
(all \(X_a\) have \(y^{+}\) labels and all \(X_d\) have \(y^{-}\) labels).
\emph{Normalized mean difference}, which also takes on values between -1 and 1,
scales these values based on the maximum possible discrimination in a dataset
given the rate of positive labels \cite{zliobaite2015survey}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.metrics import (
      mean_difference, normalized_mean_difference)

  # compare group-level discrimination in true
  # labels and predicted labels
  md_y_true = mean_difference(y, s)
  md_y_pred = mean_difference(pred, s)
  md_y_pred - md_y_true

  norm_md_y_true = norm_mean_difference(y, s)
  norm_md_y_pred = norm_mean_difference(pred, s)
  norm_md_y_pred - norm_md_y_true
\end{minted}

\cite{zliobaite2015survey} also describes \emph{consistency} and \emph{situation
test score} as individual-level discrimination measures. \emph{Consistency}
measures the difference between the target label of a particular observation and
target labels of its neighbors. K-nearest neighbors (knn) measures the pairwise
distance between observations X. Then, for each observation \(x_i\) and each
neighbor \((x_j, y_j) \in knn(x_i)\), we compute the differences between \(y_i\)
and target labels of neighbor \(y_j\). A consistency score of 0 indicates that
there is no individual-level discrimination, and a score of 1 indicates that
there is maximum discrimination in the dataset.

The \emph{situation test score} metric is similar to \emph{consistency}, except
we consider only \(x_i \in X_d\). This method uses \emph{mean difference} to
compute a discrimination score among neighbors \(x_j \in knn(x_i)\), producing a
score between 0 and 1, where 0 indicates no discrimination, and 1 indicates
maximum discrimination \cite{zliobaite2015survey}.

\begin{minted}[mathescape,
               gobble=2,
               framesep=5mm,
               fontsize=\small]{python}
  from themis_ml.metrics import (
      consistency, situation_test_score)

  # compare individual-level discrimination
  # in true labels and predicted labels
  c_true = consistency(y, s)
  c_pred = consistency(y, s)
  c_pred - c_true

  sts_true = situation_test_score(y, s)
  sts_pred = situation_test_score(y, s)
  sts_pred - sts_true
\end{minted}

\section{Evaluating Themis-ml}

In this section we use the German Credit dataset \cite{bache2013uci} to evalute
themis-ml. We use \emph{mean difference} as the ``fairness'' measure and the
\emph{area under the curve} (AUC) as the ``utility'' measure. The former
represents the degree to which PD patterns in \(D\) are learned by the ML model,
and the latter represents the predictive power of a model given the available
dataset \((X, y, s) \in D\). The following analysis is by no means meant to be a
comprehensive investigation of all possible workflows that themis- ml enables.
However, does demonstrate the potential of themis-ml as a tool that facilites
fairness-aware machine learning by enabling the user to:

\begin{enumerate}
  \item Measure PD target label distributions in the training data.
  \item Measure PD predicted labels in a machine learning algorithm's predictions.
  \item Reduce PD predictions using fairness-aware techniques.
  \item Diagnose the fairness-utility tradeoff in a particular data context.
\end{enumerate}

The German Credit dataset classifies 1000 anonymized individuals as having
``good'' and ``bad'' credit risks as part of a bank loan application, which we
encode as \(1\) and \(0\) respectively to define the \emph{credit risk}
target variable.

Each individual is associated with twenty attributes such as the purpose of the
loan, employment status, and other personal information. We begin the analysis
by extracting three protected class attributes --- \emph{female},
\emph{foreign worker}, and \emph{age below 25} --- and encode them as binary
variables such that the putatively disadvantaged group is encoded as \(1\), and
the advantaged group is encoded as \(0\) (the advantaged group would be
\emph{male}, \emph{citizen worker}, and \emph{age above 25}, respectively).

Using the \textbf{Scorer} interface, we measure PD patterns with respect to
\emph{credit risk} and each of the protected classes defined above using the
\emph{mean difference} and \emph{normalized mean difference} metrics.

\textbf{Table 3} reports the PD distribution of ``good'' and ``bad'' credit
risks with respect to the protected attributes \emph{female}, \emph{foreign
worker}, and \emph{age below 25}. The fact that both the \emph{mean difference}
(md) and \emph{normalized mean difference} (nmd) scores are greater than zero
suggests that the probability of being classified as having ``good'' risk is
higher in the advantaged group than that of the disadvantaged group.

\begin{table}
  \caption{Potentially discriminatory target variable distribution.
    \emph{md} = mean difference, \emph{nmd} = normalized mean difference.}
  \renewcommand{\arraystretch}{1.75}
  \small\noindent\begin{tabularx}{\linewidth}{l|X|X|X|X}
    \textbf{protected class} & \textbf{md (\%)} & \textbf{md 95\% CI} &
      \textbf{nmd (\%)} & \textbf{nmd 95\% CI}\\
    \hline
    \textbf{female} & 7.48 & (1.35, 13.61) & 7.73 & (1.39, 14.06) \\
    \textbf{foreign worker} & 19.93 & (4.91, 34.94) & 63.96 & (15.76, 100.00)\\
    \textbf{age below 25} & 14.94 & (7.76, 22.13) & 17.29 & (8.97, 25.61)\\
  \end{tabularx}
\end{table}

\subsection{Experimental Procedure}

To assess the extent to which (i) a model trained on these data mirrors these PD
\emph{credit risk} distributions, and (ii) fairness-aware techniques can reduce
these methods, we used \emph{mean difference} to measure model fairness and
\emph{AUC} to measure model utility. For this experiment we specify five
conditions:

\begin{itemize}
  \item Baseline (\emph{B}): Train a model on all available input variables in
        the German Credit dataset, including protected attributes.
  \item Remove Protected Attribute (\emph{RPA}): Train a model on input variables
        without protected attributes. This is the naive fairness-aware approach.
  \item Relabel Target Variable (\emph{RTV}): Train a model using the
        \emph{Relabelling} fairness-aware method.
  \item Counterfactually Fair Model (\emph{CFM}): Train a model using the
        \emph{Additive Counterfactually Fair} method.
  \item Reject-option Classification (\emph{ROC}): Train a model using the
        \emph{Reject-option Classification} method.
\end{itemize}

For each of these conditions, we train LogisticRegression, DecisionTree, and
RandomForest model types using 10-fold cross validation; generate train and test
predictions; and compute \emph{AUC} and \emph{mean difference} metrics for each
train-test pair. We then compute the mean of these metrics for each condition and
model type. The code for this analysis is available on
\underline{github} [\emph{link removed to maintain anonymity}].
% \href{https://github.com/cosmicBboy/themis-ml/blob/master/paper/Evaluating%20Themis-ml.ipynb}{\underline{github}}.

\subsection{Measuring and Mitigating Potentially Discriminatory Predictions}

Figure 1 suggests that in the case of LogisticRegression, the baseline model
\emph{B} does indeed mirror the PD patterns found in the true target variable.
Furthermore, each of the fairness-aware methods appear to have the desired
effect of reducing \emph{mean difference}, but to varying degrees depending on
the method and protected attribute. In the \emph{female} protected attribute
context, where there appears to be the least PD (mean difference of \(7.48\%\)),
the reductive effect of the fairness-aware methods do not appear to be as
large as in the \emph{foreign worker} and \emph{age below 25} contexts.

\begin{figure}[h]
\centering
\includegraphics[width=10.5cm]{fairness_aware_comparison.png}
\caption{
  \textbf{Comparison of Fairness-aware Methods} using LogisticRegression,
  DecisionTree, and RandomForest (rows) as base estimators for each protected
  attribute context (columns), measured by \emph{AUC} and
  \emph{mean difference} evaluated on test set predictions.}
\label{figure-1-comparison-fa-methods}
\end{figure}

The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA},
with respect to \emph{foreign worker} and LogisticRegression, illustrates the
observation made by \cite{kusner2017counterfactual} that removing protected
attributes from the training data does not necessarily prevent the algorithm
from mirroring PD patterns in the data.

However, the sizeable reduction in \emph{mean difference} between \emph{B} and
\emph{RPA}, with respect to \emph{age below 25} and LogisticRegression model,
shows that removing protected attributes can sometimes make models more fair
while also retaining predictive power.

An interesting thing to note here is that the \emph{Additive Counterfactually
Fair} method actually increases \emph{mean difference} for DecisionTrees and
RandomForests across all protected attribute contexts. Two possible explanations
behind this observation is that certain assumptions made by \emph{ACF} are not
suitable for non-linear learning algorithms, or the meta-estimators that compute
the residuals for non-linear estimators should be non-linear as well. This
is an open question worth future inquiry.

\subsection{The Fairness-utility Tradeoff}

Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
ML research and application \cite{fortmann2012understanding}, the fairness-
utility tradeoff can help machine learning practitioners and researchers
determine which fairness-aware methods are suitable for their particular data
context.

\begin{figure}[h]
\centering
\includegraphics[width=10.5cm]{fairness_utility_tradeoff.png}
\caption{
  \textbf{Correlation between AUC and Mean Difference} for each fairness-aware
  condition (rows) and protected attribute contexts (columns) across all model
  types (LogisticRegression, DecisionTree,   RandomForest). 95\% confidence
  intervals are provided for the pearson \(r\) correlation metric.}
\label{figure-2-fairness-utility-tradeoff}
\end{figure}

In Figure 2, we visualize the fairness-utility tradeoff, in this case as
measured by \emph{mean difference} and \emph{AUC}, respectively. We report
pearson correlation coefficients \(r\) for each protected attribute context and
fairness-aware condition with their respective 95\% confidence intervals.

These results suggest that the relationship between fairness and utility is
noisy, however there does seem to be a consistent but weak positive correlation
between \emph{mean difference} and \emph{AUC} (or a negative correlation between
fairness and utility, since lower scores are better for \emph{mean difference}
and higher scores are better for \emph{AUC}).

Interestingly, we note the cases in which there are zero or negative \(r\)
coefficient values. \(r = 0\) implies that there is no tradeoff between fairness
and utility: one can expect to increase the utility of a set of models without
adversely affecting the fairness of predictions generated by those models.
Although there are no cases where \(r_{ci\ upper} < 0\), \(r < 0\) suggests that
it might be plausible to find regimes in which one can expect to increase both
the utility and fairness of a model. Future work in this area might examine the
asymptotic behavior of the relationship between fairness and utility as model
complexity increases.

Depending on one's use cases, analyses like this might prove to be a useful
guide for figuring out what kinds of methods are robust in the sense that one
can reduce PD predictions with little to no adverse
impact on predictive performance.

\section{Discussion}

In this paper, we describe and evaluate an FMLI in the classification context
where we consider only a single binary protected class variable and a binary
target variable.

More work needs to be done to generalize FMLIs to the multi-classification,
regression, and multiple protected classes settings. Furthermore, many basic
questions about model tuning, evaluation, and selection in the fairness-aware
context remain. For instance, what might be some reasonable ways to aggregate
utility and fairness metrics in order to find the optimal set of
hyperparameters? Additionally, little is understood about the composability of
fairness-aware methods, i.e., when different techniques are used together in
sequence, are the resulting discrimination reductions additive or otherwise?

Future technical work might also extend the FMLI specification to include
techniques like Locally Interpretable Model-Agnostic Explanations
\cite{ribeiro2016should} and develop legal frameworks for thinking about how
different stakeholders would interact with FMLIs. For example, companies that
choose not to expose the model-training components of their internal ML
pipeline could still grant some form of access to the predictions generated by
the models if there were to be a set of standards for model transparency and
accountability.

Finally, many of the fairness-aware methods, such as the \emph{Relabeller},
implicitly define fairness as the uniform (equal) distribution of benefits among
disadvantaged and advantaged groups. Future work would make this definition more
flexible, for example, by defining fairness as the proportional distribution of
benefits based on need. This would necessitate the mathematical formalization of
another set of assumptions about the needs of disadvantaged and advantaged
groups.

Given the challenges ahead, our ability to measure and mitigate discrimination
is limited by our common social, legal, and political understanding of fairness
itself. This common understanding is often lacking because marginalized social
groups typically do not have a voice at the table when defining what counts as
fair. Since FMLIs are simply a tool to measure and mitigate formalized
definitions of discrimination, it is important for all stakeholders to engage in
an inclusive forum where everyone, especially disadvantaged social groups, can
contribute.

\nocite{*}
\bibliography{references}

\end{document}

### 2017-11-29 examples/measure_discrimination.py

@@ -8,7 +8,7 @@ from themis_ml.datasets import german_credit
 from themis_ml.metrics import mean_difference, normalized_mean_difference
 
 # load the german credit data
-df = german_credit()
+df = german_credit(raw=True)
 
 # target variable
 # values: 1 = low credit risk, 0 = high credit risk
@@ -32,16 +32,17 @@ foreign = df["foreign_worker"]
 # are more likely to have low credit risks compared to women and foreign
 # workers respectively.
 print("Mean difference scores:")
-print("protected class = sex: %s" % mean_difference(credit_risk, sex))
+print("protected class = sex: %s" % mean_difference(credit_risk, sex)[0])
 # 0.0748013090229
-print("protected class = foreign: %s" % mean_difference(credit_risk, foreign))
+print("protected class = foreign: %s" %
+      mean_difference(credit_risk, foreign)[0])
 # 0.199264685246
 
 print("\nNormalized mean difference scores:")
 # normalized mean difference
 print("protected class = sex: %s" %
-      normalized_mean_difference(credit_risk, sex))
+      normalized_mean_difference(credit_risk, sex)[0])
 # 0.0772946859903
 print("protected class = foreign: %s" %
-      normalized_mean_difference(credit_risk, foreign))
+      normalized_mean_difference(credit_risk, foreign)[0])
 # 0.63963963964

### 2017-11-29 paper/build-bloomberg-dx4g.sh

@@ -1,0 +1,2 @@
+pdflatex -shell-escape main
+bibtex v1_bloomberg_dx4g/main

### 2017-11-29 paper/build-j-tech-human-services.sh

@@ -1,0 +1,3 @@
+mkdir interact_output
+pdflatex -shell-escape -output-directory=interact_output main
+bibtex interact_output/main

### 2017-11-29 paper/build.sh

@@ -1,2 +1,0 @@
-pdflatex -shell-escape main
-bibtex main

### 2017-11-29 paper/deps.sh

@@ -1,0 +1,9 @@
+# deps
+tlmgr install apacite
+tlmgr install cancel
+tlmgr install floatflt
+tlmgr install morefloats
+tlmgr install multirow
+tlmgr install subfigure
+tlmgr install tabulary
+tlmgr install tfrupee

### 2017-11-29 paper/journal_tech_human_services/build.sh

@@ -1,0 +1,7 @@
+pdflatex -shell-escape main_apa6
+bibtex main_apa6
+pdflatex -shell-escape main_apa6
+
+pdflatex -shell-escape main_interact
+bibtex main_interact
+pdflatex -shell-escape main_interact

### 2018-01-07 setup.py
fix installation instructions

@@ -1,4 +1,4 @@
-from distutils.core import setup
+from setuptools import setup
 
 setup(
     name="themis-ml",
@@ -26,9 +26,9 @@ setup(
     },
     include_package_data=True,
     install_requires=[
-        "numpy",
-        "pandas",
+        "scikit-learn >= 0.19.1",
+        "numpy >= 1.9.0",
+        "pandas >= 0.22.0",
         "pathlib2",
-        "scikit-learn"
         ]
     )

### 2018-01-07 setup.py
bump to version 0.0.3

@@ -2,7 +2,7 @@ from setuptools import setup
 
 setup(
     name="themis-ml",
-    version="0.0.2",
+    version="0.0.3",
     description="Fairness-aware Machine Learning",
     author="Niels Bantilan",
     author_email="niels.bantilan@gmail.com",

### 2018-01-07 setup.py
add scipy to deps, remove include_package_data
removed the argument from `setup`, `include_package_data`,
from https://stackoverflow.com/questions/7522250/how-to-include-package-data-with-setuptools-distribute

@@ -2,7 +2,7 @@ from setuptools import setup
 
 setup(
     name="themis-ml",
-    version="0.0.3",
+    version="0.0.4",
     description="Fairness-aware Machine Learning",
     author="Niels Bantilan",
     author_email="niels.bantilan@gmail.com",
@@ -24,10 +24,10 @@ setup(
             "datasets/data/census_income_1994_1995_test.csv"
         ]
     },
-    include_package_data=True,
     install_requires=[
         "scikit-learn >= 0.19.1",
         "numpy >= 1.9.0",
+        "scipy >= 0.19.1",
         "pandas >= 0.22.0",
         "pathlib2",
         ]

### 2018-01-09 paper/journal_tech_human_services/main.tex
add corrections to themis-ml paper

--- 2017-11-29 paper/journal_tech_human_services/main_interact.tex
@@
-\documentclass{interact}
+\documentclass[man,natbib]{apa6}
 \usepackage{etoolbox}
@@
 \usepackage{hanging}
-\usepackage{hyperref}
 \usepackage{mdframed}
@@
 \usepackage{natbib}
+\usepackage{titlesec}
+\usepackage{setspace}
 \usepackage[labelfont=bf]{caption}
@@
 
-\begin{document}
+\titlespacing*{\subsection}
+{0pt}{5.5ex plus 1ex minus .2ex}{2ex plus .2ex}
+
+% code block single spacing
+\AtBeginEnvironment{minted}{\singlespacing%
+    \fontsize{10}{10}\selectfont}
 
@@
 \BeforeBeginEnvironment{minted}{
-  \begin{mdframed}[skipabove=20pt,
-                   skipbelow=10pt,
+  \begin{mdframed}[skipabove=30pt,
+                   skipbelow=20pt,
                    linewidth=0pt,
@@
 }
-
-% \author{
-%   \name{Niels Bantilan\textsuperscript{a}\thanks{Contact: Niels Bantilan. Email: niels.bantilan@gmail.com}}
-%   \affil{\textsuperscript{a}Arena.io, New York, NY}
-% }
+\shorttitle{Themis-ml: A Fairness-aware Machine Learning Interface}
 
 % For the draft version
-\author{
-  \name{Anonymous\textsuperscript{a}\thanks{Contact: anonymous. Email: anonymous}}
-  \affil{\textsuperscript{a}no affiliation}
-}
+\author{Niels Bantilan}
+\affiliation{Arena.io}
+\affiliation{New York, NY}
+\email{niels.bantilan@gmail.com}
 
-\maketitle
-
-\begin{abstract}
+\abstract{
 As more industries integrate machine learning into socially sensitive decision
@@
 learning in production systems.
-\end{abstract}
+}
 
-\begin{keywords}
-Fairness-aware machine learning; discrimination discovery; statistics;
-social bias; application programming interface
-\end{keywords}
+\keywords{
+  Fairness-aware machine learning, discrimination discovery, statistics,
+  social bias, application programming interface}
 
+\begin{document}
+
+\maketitle
 \section{Introduction} In recent years, the transformative potential of machine
@@
 methods (\citealp{kusner2017counterfactual, kamiran2012data, kamishima2012fairness,
-kamiran2012decision, zemel2013learning, zafar2017fairness, dwork2012fairness,
-zliobaite2015survey}), so we build on work done by others and seek to leverage
-these techniques in the context of research- and product-based machine learning
-applications.
+kamiran2012decision, zemel2013learning, zafar2017fairness,
+Dwork:2012:FTA:2090236.2090255, zliobaite2015survey}), so we build on work done
+by others and seek to leverage these techniques in the context of research- and
+product-based machine learning applications.
 
@@
 themis-ml, an FMLI-compliant library, and apply it to a hypothetical
-loan-granting DSS using the German Credit Dataset \cite{bache2013uci}. Finally, we
+loan-granting DSS using the German Credit Dataset (\citealp{Lichman:2013}). Finally, we
 evaluate the efficacy of themis-ml as a tool for measuring potential
@@
 credit''. While the definition above of bias is amoral, discrimination is in
-essence moral, occurring when an action is based on biases resulting in the
-systematic disenfranchisement of people based on arbitrary group membership. We
-define fairness as the inverse of discrimination, meaning that a ``fairness-
-aware'' model is one that produces non-discriminatory predictions.
+essence a moral action, occurring when biases result in the systematic
+disenfranchisement of people based on arbitrary group membership. We define
+fairness as the inverse of discrimination, meaning that a ``fairness- aware''
+model is one that produces non-discriminatory predictions.
 
@@
 determine these two types are known as disparate treatment and disparate impact,
-respectively \cite{barocas2016big}. As \cite{kusner2017counterfactual,
+respectively (\citealp{barocas2016big}). As \cite{kusner2017counterfactual,
 kamiran2012data} suggest, we can address disparate treatment in ML models by
@@
 class, and whether we are interested in measuring individual- or group-level
-discrimination \cite{dwork2012fairness}.
+discrimination (\citealp{Dwork:2012:FTA:2090236.2090255}).
 
@@
 in scikit-learn (sklearn), which is a machine learning library in the Python
-programming language \cite{scikit-learn}, and table 2 delineates some of the
+programming language (\citealp{scikit-learn}), and table 2 delineates some of the
 use cases that research or product teams might have to justify the use of an
@@
 Here we propose a high-level specification of themis-ml, an open source FMLI
-named after the ancient Greek titaness of justice (the library can be found on
+named after the ancient Greek titan of justice (the library can be found on
 % \href{https://github.com/cosmicBboy/themis-ml}{\underline{github}}.)
-\underline{github} [\emph{link removed to maintain anonymity}].)
+\underline{github} [\emph{link removed to maintain anonymity}]).
 We adopt sklearn's principles of consistency, inspection, non- proliferation of
@@
   \textbf{Model flexibility.} Focus on fairness-aware methods that are applicable
-    to a variety of model types because users might have no control or full
-    control over the specific model training implementation.
+    to a variety of model types because users have varying degrees of control
+    over the specific model training implementation.
 \end{hangparas}
@@
 \textbf{Transparency of fairness-utility tradeoff.} Fair models often make less
-  accurate predictions [8, 13], which is an important factor when assessing
-  their business impact.
+  accurate predictions (\citealp{kamishima2012fairness, zliobaite2015survey}),
+  which is an important factor when assessing their business impact.
 \end{hangparas}
@@
 interfaces. Note that we only provide a high level summary of each method,
-citing the original sources for more implementation details. The following
-descriptions make two assumptions: (i) the positive target label \(y^{+}\)
-refers to a desirable outcome, e.g. ``approve loan'', and vice versa for the
-negative target label \(y^{-}\), and (ii) the protected class is a binary
-variable defined as \(s \in \{d, a\}\), where \(X_d\)  are members of the
-disadvantaged group and \(X_a\) are members of the advantaged group.
+citing the original sources for more implementation details. Additionally, the
+themis-ml source code would also provide insights into how each method works at
+a lower level. The following descriptions make two assumptions: (i) the positive
+target label \(y^{+}\) refers to a desirable outcome, e.g. ``approve loan'', and
+vice versa for the negative target label \(y^{-}\), and (ii) the protected class
+is a binary variable defined as \(s \in \{d, a\}\), where \(X_d\)  are members
+of the disadvantaged group and \(X_a\) are members of the advantaged group.
 
@@
 \emph{Reweighting} takes a dataset D and assigns a weight to each observation
-using conditional probabilities based on \(y\) and \(s\) \cite{kamiran2012data}.
+using conditional probabilities based on \(y\) and \(s\) (\citealp{kamiran2012data}).
 In brief, large weights are assigned to \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\) ,
@@
 while removing top-ranked \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\)
-\cite{kamiran2012data}.
+(\citealp{kamiran2012data}).
 
@@
 \cite{kamishima2012fairness} proposes PRR as an optimization technique that
-extends the standard L1/L2-norm regularization method \cite{ng2004feature,
-ribeiro2016should} by adding a prejudice index term to the objective function.
-This term is equivalent to normalized mutual information, which measures the
-degree to which predictions \(y\) and \(s\) are dependent on each other. With
-values ranging from 0 to 1, 0 means that \(y\) and \(s\) are independent, and a
-value of 1 means that they are dependent. The goal of the objective function is
-to find model parameters that minimize the difference between the true label
-\(y\) and the predicted label \(\hat{y}\) in addition to the degree to which
-\(y\) depends on s.
+extends the standard L1/L2-norm regularization method (\citealp{ng2004feature,
+ribeiro2016should, park2007l1}) by adding a ``prejudice index'' term to the
+objective function. This term is equivalent to normalized mutual information,
+which measures the degree to which predictions \(y\) and \(s\) are dependent on
+each other. With values ranging from 0 to 1, 0 means that \(y\) and \(s\) are
+independent, and a value of 1 means that they are dependent. The goal of the
+objective function is to find model parameters that minimize the difference
+between the true label \(y\) and the predicted label \(\hat{y}\) in addition to
+the degree to which \(y\) depends on s.
 
@@
 \emph{Reject Option Classification} (ROC) and \emph{Discrimination Aware
-Ensemble Classification} (DAEC) \cite{kamiran2012decision}. Unlike the
+Ensemble Classification} (DAEC) (\citealp{kamiran2012decision}). Unlike the
 Transformer and Estimator methods outlined above, ROC and DAEC do not modify the
@@
 score between 0 and 1, where 0 indicates no discrimination, and 1 indicates
-maximum discrimination \cite{zliobaite2015survey}.
+maximum discrimination (\citealp{zliobaite2015survey}).
 
@@
 
-In this section we use the German Credit dataset \cite{bache2013uci} to evalute
+In this section we use the German Credit dataset (\citealp{bache2013uci}) to evaluate
 themis-ml. We use \emph{mean difference} as the ``fairness'' measure and the
@@
 dataset \((X, y, s) \in D\). The following analysis is by no means meant to be a
-comprehensive investigation of all possible workflows that themis- ml enables.
-However, does demonstrate the potential of themis-ml as a tool that facilites
+comprehensive investigation of all possible workflows that themis-ml enables.
+However, it does demonstrate the potential of themis-ml as a tool that facilites
 fairness-aware machine learning by enabling the user to:
@@
 Using the \textbf{Scorer} interface, we measure PD patterns with respect to
-\emph{credit risk} and each of the protected classes defined above using the
-\emph{mean difference} and \emph{normalized mean difference} metrics.
+\emph{credit risk} and each of the protected classes defined above using
+\emph{mean difference} and \emph{normalized mean difference}.
 
@@
 
-To assess the extent to which (i) a model trained on these data mirrors these PD
+To assess the extent to which (i) a model trained on these data mirrors PD
 \emph{credit risk} distributions, and (ii) fairness-aware techniques can reduce
-these methods, we used \emph{mean difference} to measure model fairness and
+these PD patterns, we used \emph{mean difference} to measure model fairness and
 \emph{AUC} to measure model utility. For this experiment we specify five
@@
 
-The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA},
-with respect to \emph{foreign worker} and LogisticRegression, illustrates the
-observation made by \cite{kusner2017counterfactual} that removing protected
-attributes from the training data does not necessarily prevent the algorithm
-from mirroring PD patterns in the data.
+The approximately zero reduction in \emph{mean difference} between \emph{B} and
+\emph{RPA}, with respect to \emph{foreign worker} and LogisticRegression,
+illustrates the observation made by \cite{kusner2017counterfactual} that
+removing protected attributes from the training data does not necessarily
+prevent the algorithm from mirroring PD patterns in the data.
 
@@
 \emph{RPA}, with respect to \emph{age below 25} and LogisticRegression model,
-shows that removing protected attributes can sometimes make models more fair
+shows that removing protected attributes can sometimes make models fairer
 while also retaining predictive power.
@@
 Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
-ML research and application \cite{fortmann2012understanding}, the fairness-
-utility tradeoff can help machine learning practitioners and researchers
+ML research and application \cite{fortmann2012understanding}, the fairness-utility
+tradeoff can help machine learning practitioners and researchers
 determine which fairness-aware methods are suitable for their particular data
@@
 techniques like Locally Interpretable Model-Agnostic Explanations
-\cite{ribeiro2016should} and develop legal frameworks for thinking about how
+(\citealp{ribeiro2016should}) and develop legal frameworks for thinking about how
 different stakeholders would interact with FMLIs. For example, companies that
